---
title: "A Continuous-Time Stochastic State-Space Framework for Heart Rate Variability"
subtitle: "The SDE-Inverse Gaussian Model"
author: 
  - name: "Matías Castillo-Aguilar"
    affiliation: "Escuela de Medicina, Universidad de Magallanes"
    orcid: "0000-0001-7291-247X"
bibliography: references.bib
date: today
title-block-style: manuscript
format: 
  html:
    code-fold: true
    toc: true
    number-sections: true
    fig-format: svg
---

# Abstract

Heart rate variability is widely used to probe autonomic cardiovascular regulation, yet existing models are either descriptive spectral summaries or mechanistic point-process models whose time-varying parameters lack explicit stochastic dynamics. We introduce the SDE-Inverse Gaussian (SDE-IG) framework, which couples two Ornstein-Uhlenbeck processes representing parasympathetic and sympathetic drives to an Inverse Gaussian first-passage observation model for inter-beat intervals. All ten free parameters, including the decay rates, coupling magnitudes, diffusion amplitudes, and observation parameters, are locally identifiable from a single RR interval stream: the Fisher information matrix is block-diagonal and positive definite, closed-form and profile maximum likelihood estimators exist for all parameters, and a Monte Carlo recovery study ($N = 500$ replications) confirms sub-percent relative bias and nominal 95\% confidence interval coverage. A spike-to-spike Unscented Kalman Filter recovers the net autonomic drive $\hat\Delta(t)$ and the individual branch estimates $[\hat p(t),\, \hat s(t)]$ in real time; once parameters are estimated, forward simulation of the fitted OU model generates individual branch trajectories consistent with the identified spectral structure<!-- TODO: insert real-data RMSE/KS/AIC numbers -->. The architecture extends naturally to multi-stream data, where an additional physiological observation (respiratory, blood pressure) further constrains the common-mode direction, reducing branch-level posterior uncertainty. This framework provides a principled, estimable, and interpretable generative model for beat-to-beat autonomic dynamics.

```{r}
#| label: fig-model-schematic
#| echo: false
#| fig-cap: >
#|   Generative architecture of the SDE-IG model. Two coupled Ornstein-Uhlenbeck
#|   processes represent the parasympathetic ($p(t)$, fast, HF) and sympathetic
#|   ($s(t)$, slow, LF) autonomic drives. Their difference $\Delta(t) = s(t) - p(t)$
#|   enters the log link to produce the time-varying mean interval
#|   $\mu(t) = \mu_0 \exp(-\Delta(t))$. At each moment, the elapsed time since the
#|   last beat serves as a running clock for an Inverse Gaussian first-passage
#|   hazard; a beat fires when the clock crosses a stochastic threshold. The
#|   resulting spike train is the sole observable; all ten free parameters are
#|   estimated from it via a spike-to-spike UKF (inference, right panel).

# TODO: Replace with DiagrammeR or knitr::include_graphics() once figure is created.
# Minimum viable version: grViz() call producing a left-to-right flowchart.
# Required boxes: [OU_p, OU_s] → [Δ(t)] → [log link] → [IG hazard] → [spike train]
# Required annotation: parameter labels on each component box.
knitr::asis_output("*(Figure 1 placeholder — generative architecture diagram required before submission)*")
```

# Introduction

Heart rate variability (HRV) has long been used as a noninvasive window into autonomic cardiovascular regulation. Since the early quantitative work of the 1960s and 1970s, beat-to-beat fluctuations in cardiac period have been interpreted as signatures of vagal and sympathetic modulation acting on the sinoatrial node. Classical approaches emphasized time-domain statistics and spectral decomposition, linking high-frequency oscillations to respiratory vagal activity and lower-frequency components to baroreflex-mediated dynamics. While these approaches established HRV as a core physiological measure, they are fundamentally descriptive. They characterize second-order properties of a time series without specifying a generative mechanism for heartbeat timing.

A parallel line of research sought to embed cardiac rhythm in explicit stochastic models. Early integrate-and-fire descriptions of pacemaker activity motivated first-passage-time formulations in which each heartbeat is generated when a stochastic membrane potential reaches threshold. In that context, the inverse-Gaussian distribution arises naturally as the distribution of first-passage times for a drift-diffusion process. This observation provided a physiological bridge between cellular electrophysiology and statistical modeling of inter-beat intervals. Rather than treating RR intervals as realizations of a Gaussian time series, the heartbeat sequence can be viewed as a renewal process with a parametric inter-event distribution grounded in first-passage theory.

This renewal perspective matured with the development of point-process models for HRV. In influential work by Riccardo Barbieri and colleagues [@barbieri2005point], heartbeat times were modeled as a conditional point process with inverse-Gaussian inter-event intervals whose parameters evolved in time. That framework introduced a local likelihood formulation for nonstationary RR data and established a rigorous diagnostic toolset based on the time-rescaling theorem. By transforming event times through the integrated conditional intensity, model adequacy could be assessed via standard uniformity tests, providing a principled alternative to heuristic residual analysis. These contributions repositioned HRV modeling within the broader statistical theory of event processes and made explicit the likelihood structure underlying beat-to-beat variability.

Subsequent developments extended the point-process paradigm to accommodate covariates, history dependence, and adaptive filtering [@barbieri2006special]. Time-varying parameterizations allowed heart rate to respond dynamically to respiration, blood pressure, or external perturbations, while maintaining the renewal structure of the observation model. These methods improved short-term prediction and enabled real-time tracking of instantaneous heart rate and variability. However, in many implementations, time variation was encoded through deterministic basis expansions or autoregressive updates of scalar parameters, rather than through an explicit stochastic dynamical system representing latent autonomic drives. As a result, although the models were statistically rigorous, their parameters were not always directly interpretable as physiological time constants or interaction gains.

Concurrently, state-space modeling became a central framework in systems physiology. Linear Gaussian state-space models and their nonlinear extensions were used to represent hidden regulatory processes underlying observable signals, including cardiovascular and respiratory dynamics. Stochastic differential equations, particularly the Ornstein-Uhlenbeck (OU) process, became standard tools for modeling mean-reverting physiological fluctuations. The OU process offers analytical tractability and a clear interpretation in terms of decay rates and diffusion intensity, properties that align naturally with concepts of autonomic tone and variability. In many applications, however, state-space models were applied to uniformly sampled signals, and the irregular timing of heartbeats was addressed only after interpolation or resampling.

The conceptual integration of these strands, first-passage renewal models for event timing and continuous-time stochastic dynamics for latent regulation, remains incomplete in the HRV literature. Renewal models grounded in inverse-Gaussian first-passage theory provide a physiologically motivated observation distribution, but often treat time-varying parameters as externally specified or weakly structured. State-space models capture latent dynamics, but frequently rely on Gaussian observation assumptions that are not tailored to event data. The need for models that preserve the event-based likelihood while endowing time-varying parameters with explicit stochastic dynamics is evident in studies of nonstationary protocols such as exercise onset, orthostatic challenge, or recovery, where autonomic adjustments unfold over distinct and interacting time scales.

Another unresolved issue concerns interpretability and identifiability. As HRV modeling has grown more sophisticated, the gap between statistical flexibility and physiological meaning has widened. Highly adaptive models can fit complex RR patterns but risk overparameterization and ambiguous parameter interpretation. Conversely, overly restrictive models may fail to capture essential dynamics. The current state of the art therefore demands frameworks that are both generative and constrained: generative in the sense of specifying a complete probabilistic mechanism for heartbeat production, and constrained in the sense that parameters correspond to physiologically meaningful quantities that can be interrogated and, where possible, validated.

Within this landscape, the present work positions itself as an explicit synthesis of renewal-based point-process modeling and continuous-time stochastic state dynamics. It retains the inverse-Gaussian first-passage observation structure that anchors RR intervals in integrate-and-fire theory, while introducing coupled continuous-time latent processes to represent autonomic drives with distinct time constants and potential interactions. By formalizing these latent dynamics in continuous time and linking them directly to the event-level likelihood, the model aims to extend the methodological lineage initiated by renewal and point-process approaches, rather than to replace it. The central question is whether such a synthesis can provide additional interpretive power, through identifiable dynamical parameters, without sacrificing the statistical rigor and diagnostic tools that define the event-based HRV literature.

In doing so, this work contributes to an ongoing effort to move HRV modeling beyond descriptive variability metrics toward mechanistic, testable representations of autonomic control. Its position relative to the existing literature is therefore not that of a new summary index or purely predictive algorithm, but that of a structured generative model built directly on established theoretical foundations in first-passage processes, point-process statistics, and stochastic dynamical systems.

Here we show that this synthesis is achievable and productive. We introduce the SDE-IG framework, a continuous-time stochastic state-space model coupling two Ornstein-Uhlenbeck processes to an Inverse Gaussian point process, and demonstrate that all ten free parameters are locally identifiable from a single RR interval stream, recoverable with sub-percent bias in Monte Carlo simulation, and estimable in real time via a spike-to-spike Unscented Kalman Filter. Individual autonomic branch trajectories $[\hat{p}(t), \hat{s}(t)]$ are reconstructed from the estimated parameters, providing physiologically interpretable outputs unavailable from existing point-process or spectral methods. The modular observation model extends naturally to multi-stream recordings, where additional physiological signals further constrain branch-level estimates. Taken together, these results establish the SDE-IG model as a principled, estimable, and interpretable generative framework for beat-to-beat autonomic dynamics.

# Methods

```{r}
#| label: setup
#| include: false


source("R/model_functions.R")

# ---- Shared simulation parameters ----

PARAMS <- make_model_params(
  a_p = 2.0, a_s = 0.2,          # decay rates (now free; these are sim ground truth)
  a_ps = 0.0, a_sp = 0.0,        # uncoupled baseline
  sigma_p = 0.20, sigma_s = 0.07,
  mu_0 = 1.0, rho = 0.2236068,        # rho = sqrt(mu_0/kappa): gives kappa = mu_0/rho^2 = 20
  c_p = -1.0, c_s = 0.1
)
# Derived quantity for inline use:
KAPPA_REF <- kappa_from_rho(PARAMS$free$mu_0, PARAMS$free$rho)  # = 20


INPUT_FN <- make_double_logistic(t_on = 300, t_off = 420, k = 10)

set.seed(1234)
RES <- sim_sde_ig(720, 0.005, PARAMS, INPUT_FN)

if (FALSE) {
  par(mfrow = c(1,3))
  plot(RES$time, RES$mu, type = "l")
  plot(RES$time, RES$p, type = "l")
  plot(RES$time, RES$s, type = "l")
}
```

## Latent Autonomic Dynamics

The RR interval series constrains the latent autonomic state through a single scalar observable: the net autonomic drive

$$
\Delta(t) \;=\; s(t) \;-\; p(t),
$$

which enters the mean interval via the log link. The beat-to-beat likelihood is therefore most directly informative about $\Delta(t)$, and the filter posterior contracts tightest along the differential direction $\mathbf{e}_{-} = [-1,\,1]^\top/\!\sqrt{2}$. However, this does **not** mean the individual branches are fundamentally unrecoverable. The OU parameters $\{a_p,\,a_s,\,\sigma_p,\,\sigma_s\}$ — estimated from the same RR series — fully characterize the spectral structure of each branch: $p(t)$ contributes power near $a_p/(2\pi)$ Hz (the vagal, HF pole) and $s(t)$ near $a_s/(2\pi)$ Hz (the sympathetic, LF pole). Because $a_p \gg a_s$, these poles are well-separated and individually identified from the spectral curvature of $\Delta(t)$. Two reconstruction pathways are therefore available once parameters are estimated: (1) The UKF maintains a full 2D posterior $[\hat p_k,\,\hat s_k]$ with per-branch posterior SDs; the common-mode uncertainty $\sigma_{p+s}$ is larger than the differential uncertainty $\sigma_{p-s}$ but remains bounded by the OU stationary variance. (2) Forward simulation of the fitted OU model with $\hat\Theta$ generates individual branch trajectories that are statistically consistent with the identified spectral structure and the observed $\hat\Delta(t)$.

The additive-constant non-identifiability (any constant added to both $p$ and $s$ cancels in $\Delta$) is handled by the OU prior: both branches are mean-reverting to zero, so the common mode is bounded by $\sigma_p^2/(2a_p) + \sigma_s^2/(2a_s)$. The result is that all dynamical parameters of both branches — decay rates, noise amplitudes, and coupling magnitudes — are recoverable from the single RR stream.

An additional identifiability constraint arises from the observation gains. The original formulation contained free sensitivity coefficients $k_\text{par}$ and $k_\text{sym}$ that scaled the contribution of each branch to the mean interval. Because the data identifies only $\Delta(t)$, the gains and the diffusion amplitudes $\sigma_p$, $\sigma_s$ share a common scale: multiplying $k_\text{sym}$ by $\alpha$ and dividing $s(t)$ and $\sigma_s$ by $\alpha$ leaves the likelihood unchanged for any $\alpha > 0$. This scale aliasing is resolved by normalizing $k_\text{par} = k_\text{sym} = 1$, absorbing the effective scale into $\sigma_p$ and $\sigma_s$. These are then interpretable as the amplitudes of autonomic noise in units directly linked to the log mean interval.

The resulting net drive $\Delta(t) = s(t) - p(t)$ is, under the constraints imposed here (Section 3.2), a linear combination of two independent Ornstein-Uhlenbeck processes. Its spectral density function has two poles: one in the low-frequency (LF) band determined by $a_s$, and one in the high-frequency (HF) band determined by $a_p$. This is precisely the continuous-time ARMA(2,1) [CARMA(2,1)] spectral structure that a single RR stream can identify. Its parameters are the LF and HF power amplitudes, corresponding to $\sigma_s^2/(2a_s)$ and $\sigma_p^2/(2a_p)$ respectively, together with the baseline mean interval $\mu_0$ and the shape parameter $\kappa$.

###  Structural and Free Parameter Decomposition

The system matrix is

$$
A = \begin{pmatrix} -a_p & -a_{ps} \\ -a_{sp} & -a_s \end{pmatrix}
$$

with four non-negative rate parameters. The self-decay rates $a_p$ and $a_s$ govern the intrinsic time constants of each branch ($\tau_p = 1/a_p$, $\tau_s = 1/a_s$); the cross-inhibition rates $a_{ps}$ and $a_{sp}$ encode *accentuated antagonism*, the well-documented mechanism by which sympathetic activation actively withdraws vagal tone and vice versa. All four parameters are estimated from the RR interval series; neuroscientific literature provides physiologically plausible ranges that serve as prior information for starting values ($a_p \approx 2$ Hz, $a_s \approx 0.2$ Hz) but not as fixed constraints.

Linear stability requires $\det(A) = a_p a_s - a_{ps} a_{sp} > 0$, which is enforced as a hard constraint during estimation. When the antagonism terms are zero ($a_{ps} = a_{sp} = 0$), the system reduces to two independent Ornstein-Uhlenbeck processes. Identifiability of the input sensitivities $c_p$ and $c_s$ additionally requires a non-trivial exogenous input $u(t) \not\equiv 0$. The coupling magnitudes $a_{ps}$ and $a_{sp}$ are identified from the second-order spectral structure of $\Delta(t)$: non-zero coupling shifts both eigenvalues of $A$ away from $\{-a_p, -a_s\}$, producing a detectable change in the spectral peak positions. Their separate identification from the self-decay rates requires sufficient spectral resolution, which scales with recording duration (see @sec-sensitivity).

The parameter set is therefore partitioned as follows.

```{r}
#| label: tbl-params
#| tbl-cap: "Model parameter taxonomy. Structural parameters are fixed by the neuroscience literature and are not estimated from RR data. Free parameters are identifiable from a single RR interval stream under the constraints stated in the text."
#| echo: false

param_table <- data.frame(
  Parameter = c(
    "$k_\\text{par},\\, k_\\text{sym}$",
    "$a_p$", "$a_s$",
    "$a_{ps}$", "$a_{sp}$",
    "$\\sigma_p$", "$\\sigma_s$",
    "$\\mu_0$",
    "$\\rho$",
    "$c_p,\\, c_s$"
  ),
  Class = c(
    "Structural",
    "Free", "Free", "Free", "Free",
    "Free", "Free", "Free", "Free",
    "Free (with input)"
  ),
  Value = c(
    "1 (fixed)",
    "Estimated", "Estimated",
    "Estimated ($\\geq 0$); the uncoupled model is the special case $a_{ps} = a_{sp} = 0$, tested as a constrained sub-model",
    "Estimated ($\\geq 0$); jointly identified with $a_p$ and $a_s$ from the curvature of the $\\Delta(t)$ spectral density",
    "Estimated", "Estimated",
    "Estimated", "Estimated",
    "Estimated when $u(t) \\not\\equiv 0$, else 0"
  ),
  Interpretation = c(
    "Observation gains (normalized; scale absorbed into $\\sigma$)",
    "Vagal self-decay rate; $\\tau_p = 1/a_p$",
    "Sympathetic self-decay rate; $\\tau_s = 1/a_s$",
    "Sympatho-vagal antagonism: $s(t)$ inhibits $p(t)$ (accentuated antagonism)",
    "Vagosympathetic antagonism: $p(t)$ inhibits $s(t)$ (accentuated antagonism)",
    "Vagal noise amplitude $\\to$ spectral power near $a_p/(2\\pi)$",
    "Sympathetic noise amplitude $\\to$ spectral power near $a_s/(2\\pi)$",
    "Baseline mean RR interval at autonomic neutrality",
    "$\\rho = \\sqrt{\\mu_0/\\kappa}$: baseline coefficient of variation of intervals; $\\kappa = \\mu_0/\\rho^2$",
    "Branch sensitivities to exogenous input $u(t)$"
  ),
  stringsAsFactors = FALSE
)

knitr::kable(param_table, escape = FALSE,
             col.names = c("Parameter", "Class", "Value / constraint",
                           "Physiological interpretation"))
```

###  The Net Autonomic Drive

The quantity directly estimated from the RR series is $\Delta(t) = s(t) - p(t)$. In the baseline uncoupled case ($a_{ps} = a_{sp} = 0$),

$$
d\Delta(t) \;=\; ds(t) - dp(t) \;=\;
\bigl[-a_s s(t) + c_s u(t)\bigr]dt + \sigma_s\,dW_s
-\bigl[-a_p p(t) + c_p u(t)\bigr]dt - \sigma_p\,dW_p
$$

Because $a_p \neq a_s$, $\Delta(t)$ is not itself an OU process but a difference of two OU processes with distinct time constants. When coupling is nonzero, the drift of $\Delta(t)$ involves a linear combination of both $p(t)$ and $s(t)$ individually [@sec-coupling], and $\Delta(t)$ alone is no longer sufficient to summarize the full state; the two-dimensional filter of @sec-inference is required.

The spectral density of $\Delta(t)$ is determined by the eigenvalues of $A$. Letting $\gamma_{1,2} = \tfrac{1}{2}\bigl[-(a_p + a_s) \mp \sqrt{(a_p - a_s)^2 + 4 a_{ps} a_{sp}}\bigr]$ (both negative when the stability condition holds), the spectral density takes the form of two Lorentzians centred at $|\gamma_1|/(2\pi)$ and $|\gamma_2|/(2\pi)$:

$$
S_\Delta(f) \;\approx\; \frac{A_1}{\gamma_1^2 + (2\pi f)^2}
                      + \frac{A_2}{\gamma_2^2 + (2\pi f)^2}
$$

where $A_1, A_2 > 0$ are amplitude coefficients that depend on all four rate parameters and the diffusion amplitudes. When $a_{ps} = a_{sp} = 0$, this collapses to the familiar two-Lorentzian form with poles at $a_s/(2\pi)$ and $a_p/(2\pi)$. Non-zero antagonism *spreads* the poles: $|\gamma_1|$ increases above $\max(a_p, a_s)$ and $|\gamma_2|$ decreases below $\min(a_p, a_s)$, producing a distinctive spectral signature from which all four rate parameters can, in principle, be identified. This spectrum is identifiable from the second-order properties of the RR interval series: the positions of the two spectral poles identify $a_p$ and $a_s$; the amplitudes of the two Lorentzians (proportional to $\sigma_p^2/a_p$ and $\sigma_s^2/a_s$ in the uncoupled limit; functions of all four rate parameters and both diffusion amplitudes when coupling is nonzero) encode the effective spectral power of each branch; and the non-zero coupling magnitudes $a_{ps}$ and $a_{sp}$ shift both poles away from $a_p$ and $a_s$, generating a distinctive spectral signature that distinguishes the coupled model from two independent OU processes. This directly links all ten free parameters to the spectral structure of the RR tachogram.

##  The Constrained Stochastic Differential Equations

The autonomic state vector $\mathbf{x}(t) = [p(t),\, s(t)]^\top$ evolves according to a system of coupled linear SDEs. The coupling terms $a_{ps}$ and $a_{sp}$ are free parameters estimated from data alongside $a_p$, $a_s$, $\sigma_p$, $\sigma_s$, $\mu_0$, and $\rho$. For the baseline single-stream protocol with no exogenous input, the data provide sufficient curvature to estimate the coupling magnitudes conditioned on the observed state trajectory [@sec-coupling]; the uncoupled constraint $a_{ps} = a_{sp} = 0$ is a special case used for comparison and for the initial simulation demonstrations. When coupling is nonzero, the OU log-likelihood blocks are no longer separable, and joint numerical optimization replaces the closed-form estimates [@sec-identifiability]. The full system is:

$$
\begin{aligned}
dp(t) &= \bigl[-a_p\, p(t) - a_{ps}\, s(t) + c_p\, u(t)\bigr]\,dt + \sigma_p\,dW_p(t) \\
ds(t) &= \bigl[-a_{sp}\, p(t) - a_s\, s(t) + c_s\, u(t)\bigr]\,dt + \sigma_s\,dW_s(t)
\end{aligned}
$$

where $W_p$ and $W_s$ are independent standard Wiener processes. Each branch is an Ornstein-Uhlenbeck process: mean-reverting to the homeostatic baseline (or to the input-driven equilibrium when $u(t) \neq 0$), with a time constant and noise amplitude set by its respective parameters. The free parameter set is $\{a_p,\, a_s,\, a_{ps},\, a_{sp},\, \sigma_p,\, \sigma_s,\, \mu_0,\, \rho\}$ (plus $c_p, c_s$ when input is present); all are estimated from the RR interval series. The values $a_p \approx 2\,\text{Hz}$ and $a_s \approx 0.2\,\text{Hz}$ from the autonomic neuroscience literature serve only as starting values for the optimizer; no physiological prior constrains the estimates [@sec-identifiability]. The coupling terms $a_{ps}$ and $a_{sp}$ are always free parameters; their identifiability from the RR series is established in @sec-coupling. When $a_{ps} = a_{sp} = 0$ the system reduces to two independent OU processes — a constrained sub-model used in the initial demonstrations because it admits closed-form estimators and a block-diagonal FIM that simplify exposition. When a non-trivial external input is present, $c_p$ and $c_s$ are additionally identifiable, bringing the full active set to ten parameters.

###  Vagal Kinetics and the HF Band

The parasympathetic state $p(t)$ acts as a high-pass filter on the Wiener process noise, with its spectral pole located at $a_p/(2\pi)$ Hz. The self-decay rate $a_p$ is a free parameter estimated from the RR series; values near $2$ Hz (time constant $\tau_p \approx 0.5$ s) are consistent with vagal physiology and are recovered in the simulation study below. The impulse response at these values decays within approximately $2$–$3$ s, enabling beat-to-beat vagal modulation and generating spectral power in the HF band ($0.15$–$0.4$ Hz) associated with respiratory sinus arrhythmia [@malik1996heart]. The correspondence between the estimated $\hat a_p$ and the known physiological range provides an external validity check on the parameter recovery.

```{r}
#| label: fig-impulse-response
#| fig-cap: "Theoretical impulse response of the vagal (blue) and sympathetic (red) branches. A unit impulse at $t = 1$ s illustrates the order-of-magnitude difference in recovery time. The vagal response dissipates within seconds; the sympathetic response persists for tens of seconds."
#| fig-height: 3.8

t_imp <- seq(0, 40, by = 0.05)
imp_p <- ifelse(t_imp >= 1, exp(-PARAMS$free$a_p * (t_imp - 1)), 0)
imp_s <- ifelse(t_imp >= 1, exp(-PARAMS$free$a_s * (t_imp - 1)), 0)

par(mar = c(4, 4.5, 1.5, 1))
plot(t_imp, imp_p, type = "l", col = "#0072B2", lwd = 2.5,
     xlab = "Time (s)", ylab = "Response magnitude",
     frame.plot = FALSE, ylim = c(0, 1.05))
lines(t_imp, imp_s, col = "#D55E00", lwd = 2.5)
abline(v = 1, lty = 2, col = "gray60")
tau_p_val <- round(1 / PARAMS$free$a_p, 2)
tau_s_val <- round(1 / PARAMS$free$a_s, 1)
legend("topright",
       legend = c(
         bquote("Parasympathetic" ~~ (a[p] == .(PARAMS$free$a_p) ~ Hz ~
                                      "," ~ tau[p] == .(tau_p_val) ~ s)),
         bquote("Sympathetic"     ~~ (a[s] == .(PARAMS$free$a_s) ~ Hz ~
                                      "," ~ tau[s] == .(tau_s_val) ~ s))
       ), col = c("#0072B2", "#D55E00"), lwd = 2.5, bty = "n", cex = 0.85)
grid(col = "lightgray", lty = 1)
```

###  Sympathetic Inertia and the LF Band

The sympathetic state $s(t)$ acts as a low-pass filter, with its spectral pole at $a_s/(2\pi)$ Hz. The self-decay rate $a_s$ is estimated from data; values near $0.2$ Hz (time constant $\tau_s \approx 5$ s) are consistent with noradrenergic kinetics and recovered in the simulation study. Its slow decay integrates inputs over a prolonged window, producing the smooth, trending components of autonomic regulation. The emergent spectral power is concentrated in the LF band ($0.04$–$0.15$ Hz) without any explicit oscillatory forcing, arising purely from the interaction of OU dynamics and the Wiener process.

##  Visualization of the Latent State Trajectories

Figure @fig-latent-sde shows simulated trajectories under baseline conditions, illustrating the distinct spectral signatures of the two branches and the identifiable net drive $\Delta(t) = s(t) - p(t)$.

```{r}
#| label: fig-latent-sde
#| fig-cap: "Simulated trajectories of the latent autonomic states under baseline conditions ($u(t) = 0$, all free parameters at their reference values). **Top:** Parasympathetic state $p(t)$, high-frequency texture driven by the fast decay rate. **Middle:** Sympathetic state $s(t)$, smooth, slow-trending trajectory. **Bottom:** Net autonomic drive $\\Delta(t) = s(t) - p(t)$, the quantity directly identified by the RR interval series. Positive values indicate sympathetic dominance; negative values indicate vagal dominance."
#| fig-height: 6

set.seed(101)
r0 <- sim_sde_ig(300, 0.005, PARAMS)

layout(matrix(1:3, 3, 1))
par(mar = c(1.5, 5, 1, 1.5), oma = c(3, 0, 0, 0))

plot(r0$time, r0$p, type = "l", col = "#0072B2", lwd = 1.2,
     ylab = "p(t)", xaxt = "n", frame.plot = FALSE)
abline(h = 0, lty = 2, col = "gray70"); 

plot(r0$time, r0$s, type = "l", col = "#D55E00", lwd = 2,
     ylab = "s(t)", xaxt = "n", frame.plot = FALSE)
abline(h = 0, lty = 2, col = "gray70"); 

plot(r0$time, r0$delta, type = "l", col = "#2C5F2E", lwd = 1.8,
     ylab = expression(Delta(t) == s(t) - p(t)),
     xlab = "Time (s)", frame.plot = FALSE)
abline(h = 0, lty = 2, col = "gray70"); 
mtext("Time (s)", side = 1, outer = TRUE, line = 1.5)
```


## The Inverse Gaussian Observation Model

###  Biophysical Rationale: The Integrate-and-Fire Mechanism

The mapping from continuous latent autonomic states to discrete heartbeat times is grounded in the biophysics of the sinoatrial node. During diastole, the pacemaker membrane potential rises from a hyperpolarized resting level under the influence of a net inward depolarizing current. If this potential trajectory is idealized as a Wiener process with positive drift, the time to reach the firing threshold follows an Inverse Gaussian (IG) distribution, the canonical first-passage time distribution for Brownian motion with drift [@chhikara1976optimum]. This provides a direct structural link between statistical modeling of inter-beat intervals and cardiac cellular electrophysiology, distinguishing the present framework from Poisson or Gaussian observation models that lack this mechanistic grounding.

### The Log-Link Mean Interval

The time-varying mean interval $\mu(t)$ connects the latent autonomic states to the location parameter of the IG distribution. We define this link as

$$
\mu(t) \;=\; \mu_0 \cdot \exp\!\bigl(p(t) - s(t)\bigr) \;=\; \mu_0 \cdot \exp\!\bigl(-\Delta(t)\bigr)
$$

where $\mu_0 > 0$ is the baseline mean interval at autonomic neutrality ($p = s = 0$). This **log link** replaces the linear formulation used in earlier implementations and offers three essential properties.

First, $\mu(t) > 0$ for all state values without exception, by construction. No clipping, flooring, or soft constraint is required. The linear link $\mu(t) = 1/(\rho_0 + k_\text{sym}s - k_\text{par}p)$ can produce negative or undefined values when autonomic state fluctuations push the rate below zero, requiring ad hoc correction that biases the likelihood.

Second, the log link is the canonical link for positive continuous outcomes and encodes the physiologically appropriate multiplicative structure of autonomic modulation. Autonomic tone acts multiplicatively on the depolarization drift of the SA node [@ghali2020mechanisms]; a unit increase in sympathetic tone does not add a fixed increment to heart rate but scales it by a constant factor. The log link captures this precisely: $\mu(t) = \mu_0 e^{-\Delta(t)}$, so each unit of $\Delta(t)$ scales the mean interval by $e^{-1} \approx 0.37$.

Third, the log link renders $\log\mu(t) = \log\mu_0 - \Delta(t)$ a linear function of the latent state. This is the observation equation for a Gaussian state-space model, which simplifies the linearization step in the extended Kalman filter (Phase 4) and makes the connection to existing state-space HRV methods transparent.

```{r}
#| label: fig-log-link
#| fig-cap: "Comparison of the log link and the linear link for a simulated net drive $\\Delta(t) = s(t) - p(t)$. **Left:** The linear link $\\mu(t) = 1/(\\rho_0 + \\Delta(t))$ becomes undefined and negative when $\\Delta(t)$ crosses below $-\\rho_0$, requiring clipping. **Right:** The log link $\\mu(t) = \\mu_0 \\exp(-\\Delta(t))$ is always positive and smooth. The shaded region on the left marks the invalid domain."

set.seed(123)
tt <- seq(0, 20, by = 0.02)
# Simulate a net drive with large excursions to stress-test positivity
del <- cumsum(rnorm(length(tt), 0, 0.06))
del <- del - mean(del)

rho0  <- 1 / 1.0   # baseline rate ~60 BPM
mu_ln <- 1.0 * exp(-del)
mu_li <- ifelse((rho0 + del) > 0.1, 1 / (rho0 + del), NA)

par(mfrow = c(1, 2), mar = c(4.5, 4.5, 2.5, 1))

# Left: linear link, show where it fails
ylm <- c(-0.5, max(mu_li, na.rm = TRUE) * 1.1)
plot(tt, mu_li, type = "l", col = "#D55E00", lwd = 2,
     xlab = "Time (s)", ylab = expression(mu(t) ~ "[s]"),
     main = "Linear Link", frame.plot = FALSE, ylim = ylm)
rect(xleft = 0, xright = max(tt), ybottom = -0.6, ytop = 0,
     col = rgb(0.8, 0, 0, 0.12), border = NA)
abline(h = 0, lty = 2, col = "gray50")
text(10, -0.25, "Invalid domain\n(clipping required)",
     col = "firebrick", cex = 0.8, adj = 0.5)

# Right: log link, always positive
plot(tt, mu_ln, type = "l", col = "#0072B2", lwd = 2,
     xlab = "Time (s)", ylab = expression(mu(t) ~ "[s]"),
     main = "Log Link (this model)", frame.plot = FALSE)
abline(h = 1.0, lty = 2, col = "gray50")
text(1, 0.8, expression(mu[0] == 1.0 ~ s), adj = 0, col = "gray40", cex = 0.8)
```

###  The Probability Density Function

Given that the previous beat occurred at time $u_k$, the density of the next beat time $t > u_k$ is

$$
f(t \mid u_k,\, \mathbf{x}(t)) \;=\;
\left[\frac{\kappa}{2\pi(t - u_k)^3}\right]^{1/2}
\exp\!\left\{
  -\frac{\kappa\bigl[(t-u_k) - \mu(t)\bigr]^2}{2\,\mu(t)^2\,(t-u_k)}
\right\},
$$

where $\tau = t - u_k$ is the elapsed time since the last beat and $\kappa = \mu_0/\rho^2 > 0$ is the derived shape parameter and $\rho = \sqrt{\mu_0/\kappa}$ is the baseline coefficient of variation estimated directly from the data. In the first-passage interpretation, $\rho$ is proportional to the threshold-to-noise ratio of the SA node membrane: small $\rho$ corresponds to a near-deterministic rhythm; large $\rho$ to a highly variable, noise-dominated one. Typical resting values are $\rho \approx 0.15$--$0.30$.

The shape parameter $\kappa > 0$ controls interval regularity but is not directly interpretable in variance units. We therefore work with the baseline coefficient of variation

$$
\rho \;=\; \sqrt{\mu_0/\kappa}
$$

so that $\kappa = \mu_0/\rho^2$ and $\operatorname{Var}(\tau_0) = \rho^2 \mu_0^2$ at autonomic neutrality ($\Delta = 0$). For a general state, $\operatorname{Var}(\tau) = \rho^2 \mu(t)^3/\mu_0$, preserving the heteroscedastic property while making the baseline variability directly readable: $\rho \approx 0.22$ corresponds to a coefficient of variation of 22\%, typical of resting HRV. The free parameter estimated from data is $\rho$; all likelihood computations use $\kappa = \mu_0/\rho^2$ internally.

A distinctive property of the IG distribution is its **heteroscedastic variance**: $\text{Var}(\tau) = \mu(t)^3/\kappa$. The cubic dependence on the mean implies that as heart rate increases (and $\mu$ decreases), beat-to-beat variability decreases automatically, even when $\kappa$ is held constant. This provides a mechanistic account of the empirically well-documented reduction in HRV during exercise and stress, without requiring any model-specific adjustments.

###  Log-Space Hazard Computation

For the conditional intensity function, the instantaneous probability of a beat at time $t$ given survival to time $t$, we require the hazard rate $\lambda(t \mid \mathcal{H}_t) = f(\tau)/S(\tau)$, where $S(\tau) = 1 - F(\tau)$ is the survival function. The IG CDF is

$$
F(\tau;\mu,\kappa) \;=\;
\Phi\!\left(\sqrt{\tfrac{\kappa}{\tau}}\Bigl(\tfrac{\tau}{\mu}-1\Bigr)\right)
+
\exp\!\left(\tfrac{2\kappa}{\mu}\right)
\Phi\!\left(-\sqrt{\tfrac{\kappa}{\tau}}\Bigl(\tfrac{\tau}{\mu}+1\Bigr)\right)
$$

The term $\exp(2\kappa/\mu)$ grows rapidly: at $\kappa = 15$, $\mu = 0.5$ s, its value is $e^{60} \approx 10^{26}$. Although this is within IEEE 754 double-precision range, the numerical product $\exp(2\kappa/\mu) \cdot \Phi(-z_2)$ involves multiplying a huge number by a tiny one, causing catastrophic cancellation that propagates into both the survival function and the hazard. More critically, any gradient-based optimization of the likelihood will encounter numerical instability at precisely the parameter values of greatest interest, those where heart rate is elevated.

The resolution is to compute the log-survival entirely in log space. Defining

$$
\log A \;=\; \log\Phi(-z_1), \qquad
\log B \;=\; \tfrac{2\kappa}{\mu} + \log\Phi(-z_2)
$$

the log-survival is

$$
\log S(\tau) \;=\; \log A + \log\!\bigl(1 - e^{\,\log B - \log A}\bigr)
              \;=\; \log A + \texttt{log1mexp}(d)
$$

where $d = \log B - \log A < 0$ in the physiological range and `log1mexp` is the numerically stable evaluation of $\log(1 - e^x)$ for $x < 0$:

$$
\texttt{log1mexp}(x) \;=\;
\begin{cases}
\log(-\text{expm1}(x)) & x \geq \log(0.5) \\
\text{log1p}(-\exp(x)) & x < \log(0.5)
\end{cases}
$$

The two branches use the most numerically stable form for each range, avoiding catastrophic cancellation uniformly. The log-hazard is then simply $\log\lambda(\tau) = \log f(\tau) - \log S(\tau)$, computed entirely in log space with no intermediate overflow.

```{r}
#| label: fig-ig-shapes
#| fig-cap: >
#|   The Inverse Gaussian probability density. **Left:** Effect of the 
#|   mean interval $\\mu$ (shape fixed at $\\kappa = 15$). As $\\mu$ decreases 
#|   (sympathetic dominance), the distribution shifts to earlier times. 
#|   **Right:** Effect of the shape parameter $\\kappa$ (mean fixed at 
#|   $\\mu = 1.00$ s). Higher $\\kappa$ corresponds to lower variance, 
#|   a more regular rhythm. The heteroscedastic property 
#|   $\\text{Var}(\\tau) = \\mu^3/\\kappa$ links both panels: 
#|   reducing $\\mu$ reduces the variance even at fixed $\\kappa$.
#|   These are illustrative parameter values chosen to show the shape of the distribution
#|   across a physiological range; the reference simulation uses $\\mu_0 = `r PARAMS$free$mu_0`$ s
#|   and $\\kappa = `r round(kappa_from_rho(PARAMS$free$mu_0, PARAMS$free$rho))`$.
#| fig-height: 4

t_sq  <- seq(0.01, 2.2, length.out = 400)
dig_v <- function(t, mu, k) exp(log_ig_pdf(t, mu, k))

mu_v  <- c(0.5, 0.85, 1.2)
k_v   <- c(5, 15, 50)
cols  <- c("#D55E00", "black", "#0072B2")

par(mfrow = c(1, 2), mar = c(4.5, 4.5, 2.5, 1))

# Left: varying mu
plot(NA, xlim = c(0, 2.2), ylim = c(0, 6),
     xlab = "Interval (s)", ylab = "Density",
     main = "Effect of mean interval \u03bc", frame.plot = FALSE)
for (i in 1:3) lines(t_sq, dig_v(t_sq, mu_v[i], 15), col = cols[i], lwd = 2)
legend("topright",
       legend = paste("\u03bc =", mu_v, "s"),
       col = cols, lwd = 2, bty = "n", cex = 0.85)

# Right: varying kappa
plot(NA, xlim = c(0, 2.2), ylim = c(0, 6),
     xlab = "Interval (s)", ylab = "Density",
     main = "Effect of shape \u03ba", frame.plot = FALSE)
for (i in 1:3) lines(t_sq, dig_v(t_sq, 0.85, k_v[i]), col = cols[i], lwd = 2)
legend("topright",
       legend = paste("\u03ba =", k_v),
       col = cols, lwd = 2, bty = "n", cex = 0.85)
```

###  The Hazard Function and Its Physiological Significance

The resulting hazard function $\lambda(\tau)$ is non-monotonic. At $\tau = 0$, the hazard is zero, correctly modeling the absolute refractory period of cardiac tissue. It rises steeply as $\tau$ approaches $\mu(t)$, the membrane potential nears threshold, then peaks and gradually decreases in the tail. Under the log link, a shift in $\Delta(t)$ compresses or stretches this profile continuously, modulating the probability of an earlier or later beat without discontinuity.

```{r}
#| label: fig-hazard-function
#| fig-cap: "The Inverse Gaussian hazard function $\\lambda(\\tau \\mid \\mathcal{H}_t)$, computed using the fully log-space algorithm. **Left:** Effect of sympathetic activation: reducing $\\mu(t)$ shifts the hazard peak to the left, increasing the probability of an earlier beat. **Right:** The hazard surface as a function of both elapsed time and the current mean interval. Brighter colors indicate higher instantaneous firing probability."

tau_v  <- seq(0.01, 2.5, length.out = 500)

kap_fig <- kappa_from_rho(PARAMS$free$mu_0, PARAMS$free$rho)
mu_base <- PARAMS$free$mu_0

h_base  <- vapply(tau_v, ig_hazard, numeric(1), mu = mu_base,        kappa = kap_fig)
h_sym   <- vapply(tau_v, ig_hazard, numeric(1), mu = mu_base * 0.60, kappa = kap_fig)
h_para  <- vapply(tau_v, ig_hazard, numeric(1), mu = mu_base * 1.40, kappa = kap_fig)

mu_ax  <- seq(mu_base * 0.4, mu_base * 1.5, length.out = 60)
tau_ax <- seq(0.02, 1.8, length.out = 60)
hgrid  <- outer(tau_ax, mu_ax, Vectorize(function(ta, mu) ig_hazard(ta, mu, kap_fig)))

par(mfrow = c(1, 2), mar = c(4.5, 5, 2.5, 1))

plot(tau_v, h_base, type = "l", lwd = 2.5, col = "black",
     ylim = c(0, max(c(h_base, h_sym, h_para)) * 1.08),
     xlab = "Time since last beat (s)",
     ylab = expression(lambda(tau ~ "|" ~ H[t])),
     main = "Hazard Function", frame.plot = FALSE)
lines(tau_v, h_sym,  lwd = 2.5, col = "#D55E00", lty = 2)
lines(tau_v, h_para, lwd = 2.5, col = "#0072B2", lty = 3)
legend("topright",
       legend = c(bquote(mu == .(mu_base) ~ s ~ "(baseline)"),
                  bquote(mu == .(round(mu_base * 0.60, 2)) ~ s ~ "(sympathetic \u2191)"),
                  bquote(mu == .(round(mu_base * 1.40, 2)) ~ s ~ "(vagal \u2191)")),
       col = c("black", "#D55E00", "#0072B2"),
       lwd = 2.5, lty = c(1, 2, 3), bty = "n", cex = 0.85)

image(tau_ax, mu_ax, hgrid,
      col  = hcl.colors(100, "YlOrRd", rev = TRUE),
      xlab = "Time since last beat (s)",
      ylab = expression("Mean interval" ~ mu(t) ~ "[s]"),
      main = "Hazard Surface", axes = FALSE)
axis(1); axis(2)
contour(tau_ax, mu_ax, hgrid, add = TRUE,
        col = "black", nlevels = 5, lwd = 0.8)
```

## Numerical Simulation Algorithm

The computational implementation of the framework requires a hybrid scheme that bridges continuous-time latent dynamics and discrete event generation. Three structural improvements over naive implementations are incorporated: an exact discretization of the OU transition, a fully log-space intensity evaluation, and a survival-ratio event generation procedure that avoids the systematic bias of the Bernoulli approximation.

###  Exact Ornstein-Uhlenbeck Transition Kernel

The OU process $dx = (-ax + cu)\,dt + \sigma\,dW$ has an exact Gaussian transition density. Because the drift is linear in the state, the Itô solution can be computed in closed form: for any $\Delta t > 0$,

$$
x(t + \Delta t) \mid x(t) \;\sim\;
\mathcal{N}\!\left(
  x(t)\,e^{-a\Delta t} + \tfrac{cu}{a}(1 - e^{-a\Delta t}),\quad
  \tfrac{\sigma^2}{2a}\,(1 - e^{-2a\Delta t})
\right)
$$

This transition is exact at any step size. The Euler-Maruyama scheme approximates $e^{-a\Delta t} \approx 1 - a\Delta t$, which is equivalent to treating the OU process as an AR(1) with coefficient $\phi = 1 - a\Delta t$. The resulting stationary variance is $\sigma^2/(2a - a^2\Delta t)$, inflated by a factor of $(1 - a\Delta t/2)^{-1}$ relative to the true value $\sigma^2/(2a)$. For the vagal branch ($a_p = 2.0$, $\Delta t = 0.005$ s) this is a $0.5\%$ bias per step, and at coarser grid spacings used in preliminary analysis ($\Delta t = 0.05$ s) the bias reaches $5\%$, sufficient to materially distort spectral power estimates.

Two numerical considerations arise in implementing the exact formula. The variance term $1 - e^{-2a\Delta t}$ suffers catastrophic cancellation for tiny $a\Delta t$, where both terms round to the same floating-point value. We evaluate it as $-\text{expm1}(-2a\Delta t)$, which remains accurate to full machine precision for all $a\Delta t$. Similarly, the mean shift $1 - e^{-a\Delta t}$ is evaluated as $-\text{expm1}(-a\Delta t)$.

```{r}
#| label: fig-ou-exact
#| cache: true
#| fig-cap: "Stationary variance bias of Euler-Maruyama (EM) versus the exact OU kernel, shown for the vagal branch ($a_p = 2.0$ Hz, $\\sigma_p = 0.20$). **Left:** Empirical stationary variance from 300 replicate trajectories ($N = 10000$ steps each); The exact kernel achieves the true value $\\sigma^2/(2a) = 0.01$ at every step size. **Right:** Relative bias (%) as a function of step size, confirming the theoretical prediction $\\approx a\\Delta t/2$."
#| fig-height: 4

set.seed(321)
a_p  <- PARAMS$free$a_p;  sig_p <- PARAMS$free$sigma_p
true_var <- sig_p^2 / (2 * a_p)
dt_vals  <- c(0.005, 0.01, 0.025, 0.05, 0.1)
nrep     <- 300;  nstep <- 5000;  burn <- 1000

em_var <- exact_var <- numeric(length(dt_vals))

for (j in seq_along(dt_vals)) {
  dt <- dt_vals[j]
  em_v <- exact_v <- numeric(nrep)
  for (r in seq_len(nrep)) {
    xe2  <- 0; xex2 <- 0
    xe_traj  <- numeric(nstep)
    xex_traj <- numeric(nstep)
    for (i in seq_len(nstep)) {
      xe2  <- xe2  + (-a_p * xe2  * dt + sig_p * sqrt(dt) * rnorm(1))
      xex2 <- ou_exact_step(xex2, a_p, sig_p, dt = dt)
      xe_traj[i] <- xe2; xex_traj[i] <- xex2
    }
    em_v[r]    <- var(xe_traj[(burn+1):nstep])
    exact_v[r] <- var(xex_traj[(burn+1):nstep])
  }
  em_var[j]    <- mean(em_v)
  exact_var[j] <- mean(exact_v)
}

par(mfrow = c(1, 2), mar = c(4.5, 5, 2.5, 1))

# Left: absolute variance
plot(dt_vals * 1000, em_var,    type = "b", pch = 19, lwd = 2,
     col = "#D55E00", ylim = c(true_var * 0.95, max(em_var) * 1.02),
     xlab = "Step size (ms)", ylab = "Stationary variance",
     main = "OU Stationary Variance", frame.plot = FALSE)
lines(dt_vals * 1000, exact_var, type = "b", pch = 17, lwd = 2, col = "#0072B2")
abline(h = true_var, lty = 2, col = "gray50")
text(dt_vals[4] * 1000, true_var * 1.003,
     bquote(sigma^2 / (2*a) == .(round(true_var, 4))),
     adj = 0, cex = 0.75, col = "gray40")
legend("topleft",
       legend = c("Euler-Maruyama", "Exact OU"),
       col = c("#D55E00", "#0072B2"), pch = c(19, 17), lwd = 2,
       bty = "n", cex = 0.85)

# Right: relative bias
em_bias    <- 100 * (em_var    / true_var - 1)
exact_bias <- 100 * (exact_var / true_var - 1)
plot(dt_vals * 1000, em_bias, type = "b", pch = 19, lwd = 2, col = "#D55E00",
     xlab = "Step size (ms)", ylab = "Relative bias (%)",
     main = "Discretization Bias", frame.plot = FALSE,
     ylim = range(c(em_bias, exact_bias)))
lines(dt_vals * 1000, exact_bias, type = "b", pch = 17, lwd = 2, col = "#0072B2")
abline(h = 0, lty = 2, col = "gray50")
```

###  Log-Space Intensity Evaluation

At each simulation step, the current OU state is mapped to the mean interval via $\mu(t) = \mu_0 \exp(p(t) - s(t))$, and the conditional intensity is evaluated as $\log\lambda(\tau) = \log f(\tau;\mu,\kappa) - \log S(\tau;\mu,\kappa)$, following the log-space derivation in Section 4.4. Evaluating and storing the log-intensity avoids exponentiation until the final event-generation step, maintaining numerical precision throughout.

###  Survival-Ratio Event Generation

The Bernoulli approximation $P(\text{event in } [t_i, t_{i+1}]) \approx \lambda(t_i)\Delta t$ produces a signed error whose direction depends on the slope of the hazard at the step entry point. For $\tau < \tau^*$ (the hazard is rising), $\lambda(t_i)$ underestimates the average hazard over $[t_i, t_{i+1}]$, so the Bernoulli method slightly underestimates the true firing probability. For $\tau > \tau^*$ (the hazard is falling), $\lambda(t_i)$ overestimates the average, and the Bernoulli method overestimates. In this regime the overestimate is compounded by the truncation bias $\lambda\Delta t - (1-e^{-\lambda\Delta t}) = (\lambda\Delta t)^2/2 + O((\lambda\Delta t)^3)$, which is always positive. Near the hazard peak, where $\lambda \approx 15$–$25$ Hz and $\Delta t = 0.005$ s, the overestimate reaches 3–6%, a directional error that inflates event counts and compresses estimated inter-beat intervals. Because the IG density mode lies just before the hazard peak, the majority of beats are generated in the overestimating regime, producing a net positive bias in event rate across a full simulation.

The exact within-step firing probability, under the assumption that $\mu(t)$ is constant within the interval $[t_i, t_{i+1}]$, is obtained directly from the survival function ratio:

$$
P\!\bigl(\text{event in } [t_i, t_{i+1})\bigr)
\;=\; 1 - \frac{S(\tau + \Delta t;\,\mu_i,\,\kappa)}{S(\tau;\,\mu_i,\,\kappa)}
$$

In log space:

$$
P \;=\; -\text{expm1}\!\bigl(\log S(\tau + \Delta t) - \log S(\tau)\bigr)
$$

using the identity $1 - e^x = -\text{expm1}(x)$, which is numerically stable for all $x \leq 0$. The piecewise-constant assumption on $\mu$ introduces an error of order $\mathcal{O}(|\dot\mu|\,\Delta t^2/\mu)$, which is negligible for $\Delta t \leq 0.005$ s given the smooth OU dynamics.

```{r}
#| label: fig-bernoulli-vs-survival
#| fig-cap: "Systematic bias of the Bernoulli approximation relative to the exact survival-ratio method. **Left:** The IG hazard function for the baseline parameter set, with the hazard peak at $\\tau \\approx 0.72$ s. **Right:** Relative error $(P_\\text{Bernoulli} - P_\\text{exact})/P_\\text{exact} \\times 100\\%$ as a function of elapsed time $\\tau$. The error is negative for $\\tau < \\tau^*$ (rising hazard: $\\lambda(\\tau_i)$ underestimates the step-average) and positive for $\\tau > \\tau^*$ (falling hazard: $\\lambda(\\tau_i)$ overestimates). The maximum positive overestimate, annotated on the right panel, occurs just past the hazard peak where $\\lambda$ is largest and the surplus $\\lambda(\\tau_i)\\Delta t - P_{\\text{exact}}$ peaks. Because most beats are generated in this post-peak regime, the net bias across a full simulation is a systematic overestimate of event counts."

fs <- study_firing_bias(mu = PARAMS$free$mu_0, 
                        kappa = kappa_from_rho(PARAMS$free$mu_0, PARAMS$free$rho),
                        dt = 0.005, 
                        tau_range = c(0.3, 3 * PARAMS$free$mu_0))

par(mfrow = c(1, 2), mar = c(4.5, 5, 2.5, 1))

plot(fs$tau, fs$lambda, type = "l", lwd = 2.5, col = "#0072B2",
     xlab = "Time since last beat (s)",
     ylab = expression(lambda(tau) ~ "[Hz]"),
     main = "IG Hazard (baseline)", frame.plot = FALSE)
pk  <- fs$tau[which.max(fs$lambda)]
lmx <- max(fs$lambda)
abline(v = pk, lty = 2, col = "gray50")
text(pk + 0.04, lmx * 0.95, sprintf("\u03c4* = %.2f s", pk),
     adj = 0, cex = 0.8, col = "gray40")

plot(fs$tau, fs$rel_error_pct, type = "l", lwd = 2.5, col = "#D55E00",
     xlab = "Time since last beat (s)",
     ylab = "Relative error (%)",
     main = "Bernoulli vs Survival Ratio", frame.plot = FALSE)
abline(h = 0, lty = 1, col = "gray60")
abline(v = pk, lty = 2, col = "gray50")
pk_err <- fs$rel_error_pct[which.max((fs$rel_error_pct))]
pk_tau <- fs$tau[which.max((fs$rel_error_pct))]
points(pk_tau, pk_err, pch = 19, col = "#D55E00", cex = 1.5)
text(pk_tau + 0.06, pk_err,
     sprintf("Peak overestimate: +%.2f%%", pk_err), adj = 0, cex = 0.8, col = "#D55E00")
```

###  Exogenous Input Protocol

The simulation study uses a **double-logistic** input function to represent a rest–exercise–recovery protocol:

$$
u(t) \;=\; \sigma\!\bigl(k(t - t_{\rm on})\bigr)\cdot\sigma\!\bigl(k(t_{\rm off} - t)\bigr),
\qquad \sigma(x) = (1 + e^{-x})^{-1}
$$

with $t_{\rm on} = 300$ s, $t_{\rm off} = 420$ s, and steepness $k = 10$ Hz. At $k = 10$ the 10–90\% rise time is $\ln(81)/k \approx 0.44$ s, which is negligible relative to both the vagal time constant ($\tau_p = 0.5$ s) and the sympathetic time constant ($\tau_s = 5$ s), so the transient is physiologically abrupt while remaining smooth and differentiable at all $t$. The boxcar $\mathbf{1}_{[t_{\rm on}, t_{\rm off})}(t)$ is the $k \to \infty$ limit; its breakpoint discontinuities create non-smooth gradients in the OU input-driven mean shift, which impairs profile likelihood computations and MLE convergence. The double-logistic eliminates this artifact without altering the qualitative autonomic response.

###  Complete Simulation Algorithm

The full simulation integrates the three components above into a single time-indexed loop. At each step $i$:

1. **State update**: draw $p_{i+1}$ and $s_{i+1}$ from the exact OU transition densities with decay rates $a_p$, $a_s$ and diffusion amplitudes $\sigma_p$, $\sigma_s$.
2. **Link**: compute $\mu_{i+1} = \mu_0\exp(p_{i+1} - s_{i+1})$ and $\Delta_{i+1} = s_{i+1} - p_{i+1}$.
3. **Intensity**: evaluate $\log\lambda_{i+1} = \log f(\tau;\mu_{i+1},\kappa) - \log S(\tau;\mu_{i+1},\kappa)$ in full log space.
4. **Event**: compute $P = -\text{expm1}(\log S(\tau + \Delta t) - \log S(\tau))$; draw $U \sim \text{Uniform}(0,1)$; if $U < P$, record a spike and reset $\tau \leftarrow 0$.

# Identifiability Analysis {#sec-identifiability}

The framework introduces up to ten free parameters: $\Theta = \{a_p,\, a_s,\, a_{ps},\, a_{sp},\, \sigma_p,\, \sigma_s,\, \mu_0,\, \rho,\, c_p,\, c_s\}$. The input sensitivities $c_p$ and $c_s$ are identifiable only when $u(t) \not\equiv 0$. The coupling terms $a_{ps}$ and $a_{sp}$ are always free parameters; their conditional identifiability from a single RR stream is established in @sec-coupling. For the baseline uncoupled analysis ($a_{ps} = a_{sp} = 0$) the active set comprises six always-free parameters $\{a_p, a_s, \sigma_p, \sigma_s, \mu_0, \rho\}$ plus two protocol-conditional parameters $\{c_p, c_s\}$, identifiable only when $u(t) \not\equiv 0$; the stress protocol used here activates all eight.  The FIM analysis below therefore covers the full $8 \times 8$ system. Before applying the model to data, we must establish that the parameters are identifiable and recoverable with quantified precision.  We address this in two stages. The present section establishes conditional identifiability: given the latent state trajectory $\{p(t), s(t)\}_{0 \le t \le T}$, the full parameter set $\Theta$ is recoverable from the resulting beat times with sub-percent bias and nominal confidence interval coverage. Three interlocking demonstrations are provided: an analytical argument based on log-likelihood separability, a numerical eigenspectrum of the observed FIM, and a Monte Carlo recovery study ($N = 500$ replications). For the uncoupled sub-model ($a_{ps} = a_{sp} = 0$) with a non-trivial input ($u(t) \not\equiv 0$), the FIM is $8 \times 8$ and block-diagonal — this separability yields closed-form estimators for each block, which we exploit for the primary demonstrations. Identifiability of the coupling terms in the full $10 \times 10$ system, where the block-diagonal structure breaks down, is established in @sec-coupling via a numerical FIM and a dedicated recovery study. Marginal identifiability — recovery from the spike train alone, integrating over unobserved latent states — is established in @sec-inference via the UKF marginal likelihood.

```{r}
#| label: ident-setup
#| include: false
source("R/mle.R")
source("R/identifiability.R")
```

## Separability of the Conditional Log-Likelihood

For the uncoupled sub-model ($a_{ps} = a_{sp} = 0$), the conditional log-likelihood separates into three independent blocks:


$$
\log \mathcal{L}(\Theta \mid \mathbf{x}_{0:T},\, \{t_k\})
= \underbrace{\log \mathcal{L}_p(a_p,\, \sigma_p,\, c_p)}_{\text{OU parasympathetic}}
+ \underbrace{\log \mathcal{L}_s(a_s,\, \sigma_s,\, c_s)}_{\text{OU sympathetic}}
+ \underbrace{\log \mathcal{L}_\text{obs}(\mu_0,\, \rho)}_{\text{IG observations}}
$$

where the three components arise from the OU transition densities of $p(t)$ and $s(t)$ and from the IG beat-time densities respectively. 

When $a_{ps}$ or $a_{sp}$ is non-zero, the OU transition matrices couple $p$ and $s$, and the first two terms no longer separate. In that case the OU log-likelihood must be maximized jointly over all eight OU parameters numerically (see @sec-coupling). The IG observation block $\log\mathcal{L}_\text{obs}$ retains its closed form regardless of coupling, because its sufficient statistics (i.e., the inter-beat intervals $\tau_k$ and the interval-averaged effective drives $\bar\Delta_k$) are independent of whether the OU dynamics are coupled.

This separability has two important consequences.

**Consequence 1: closed-form MLEs.** Each block admits an analytic maximum. For the OU blocks, the data generating structure is a Gaussian linear model with known design matrix:

$$
x_{i+1} = x_i e^{-a\Delta t} + \frac{c\, u_i}{a}(1 - e^{-a\Delta t}) + \varepsilon_i, \qquad
\varepsilon_i \sim \mathcal{N}\!\left(0,\, \frac{\sigma^2}{2a}(1 - e^{-2a\Delta t})\right)
$$

The decay rate $a$ is first estimated by one-dimensional profile likelihood over $\log a$, with $(\hat{c}, \hat{\sigma})$ computed in closed form at each candidate value: $\hat{c} = \sum z_i y_i / \sum z_i^2$ (OLS) and $\hat\sigma^2 = \overline{r_i^2}/C_a$, where $r_i$ are OLS residuals and $C_a = (1 - e^{-2a\Delta t})/(2a)$. For the IG observation block, the score equations yield:

$$
\hat\mu_0 = \frac{\sum_k \tau_k\, e^{2\bar\Delta_k}}{\sum_k e^{\bar\Delta_k}}, \qquad
\hat\kappa = \frac{N}{\displaystyle\sum_k \frac{(\tau_k - \hat\mu_k)^2}{\hat\mu_k^2\,\tau_k}}
$$

where $\hat\mu_k = \hat\mu_0 e^{-\bar\Delta_k}$, and $\hat\rho = \sqrt{\hat\mu_0/\hat\kappa}$ is recovered by the delta method.  Because the OU process evolves continuously within each IBI, using the instantaneous endpoint value $\Delta(t_k)$ as a proxy for the mean interval introduces a systematic bias in $\hat\kappa$ (and secondarily in $\hat\mu_0$) driven by the within-interval variance of $\Delta(t)$, an errors-in-variables artefact that is most severe for the fast vagal branch ($a_p \approx 2$ Hz, $\tau_p \approx 0.5$ s, comparable to a typical IBI). The estimator therefore uses the **interval-averaged effective drive**

$$
\bar\Delta_k \;=\; -\log\!\left(\frac{1}{n_k}\sum_{j \in \text{IBI}_k} e^{-\Delta(t_j)}\right)
$$

computed from the stored latent trajectory over the $n_k$ grid points interior to $(t_{k-1}, t_k]$. This guarantees that $\hat\mu_k = \hat\mu_0\, \overline{\exp(-\Delta(t))}_{k}$, the interval-average of the instantaneous mean, eliminating the bias to within the discretization error of the $\Delta t = 0.005$ s grid (confirmed in the recovery study below).

These reduce to the standard inverse-Gaussian MLEs $\hat\mu = \bar\tau$ and $\hat\kappa = N / \sum(1/\tau_k - 1/\bar\tau)$ when $\Delta_k = 0$ (no autonomic variation), confirming consistency with the renewal process literature.

**Consequence 2: block-diagonal FIM.** The Fisher information matrix is block-diagonal in the parameterization $(\log a_p,\, \log\sigma_p,\, c_p,\, \log a_s,\, \log\sigma_s,\, c_s,\, \log\mu_0,\, \log\rho)$. Identifiability reduces to showing that each block is positive definite: the two $3 \times 3$ OU blocks $({\log a_p,\, \log\sigma_p,\, c_p})$ and $({\log a_s,\, \log\sigma_s,\, c_s})$ and the $2 \times 2$ observation block. This block-diagonal structure holds for $a_{ps} = a_{sp} = 0$; when coupling is nonzero the blocks link and the FIM is computed as a single $10 \times 10$ numerical object over $({\log a_p, \log a_s, a_{ps}, a_{sp}, \log\sigma_p, \log\sigma_s, c_p, c_s, \log\mu_0, \log\rho})$ [@sec-coupling].

## Observed Fisher Information Matrix

```{r}
#| label: fig-fim
#| fig-cap: "Observed Fisher information matrix and eigenvalue spectrum for the reference parameter set. **Left:** The normalized FIM (each entry divided by the maximum diagonal value) displayed as a heatmap. The three blocks along the diagonal correspond to the parasympathetic OU block $(\\log a_p,\\, \\log\\sigma_p,\\, c_p)$, sympathetic OU block $(\\log a_s,\\, \\log\\sigma_s,\\, c_s)$, and IG observation block $(\\log\\mu_0,\\, \\log\\rho)$. Off-diagonal blocks are exactly zero by construction, confirming the theoretical separability of the three log-likelihood components. **Right:** Eigenvalues on a log scale. All eight eigenvalues are strictly positive, confirming that the FIM is positive definite and all parameters are locally identified. The large spread reflects the differing amounts of information available from $n \\approx 60{,}000$ OU transitions versus $N \\approx 750$ beats."
#| fig-height: 4.5

fim_result <- full_conditional_fim(RES)

par(mfrow = c(1, 2), mar = c(5, 6, 2.5, 1))

# Left: normalized FIM heatmap
F_norm <- abs(fim_result$FIM) / max(abs(diag(fim_result$FIM)))
np_fim <- length(fim_result$param_names)
image(seq_len(np_fim), seq_len(np_fim), t(F_norm[np_fim:1, ]),
      col  = hcl.colors(64, "Blues", rev = TRUE),
      xlab = "", ylab = "",
      axes = FALSE,
      main = "Normalized FIM")
axis(1, at = seq_len(np_fim), labels = fim_result$param_names, las = 2, cex.axis = 0.75)
axis(2, at = seq_len(np_fim), labels = rev(fim_result$param_names), las = 2, cex.axis = 0.75)
box()

# Right: eigenvalue spectrum
eig_sorted <- sort(fim_result$eigenvalues, decreasing = TRUE)
barplot(log10(eig_sorted),
        names.arg = paste0("\u03bb", seq_along(eig_sorted)),
        col = "#0072B2",
        xlab = "Eigenvalue index",
        ylab = expression(log[10](lambda[i])),
        main = "FIM Eigenvalue Spectrum",
        border = NA)
abline(h = 0, lty = 2, col = "gray50")
grid(nx = NA, col = "lightgray", lty = 1)
```

```{r}
#| label: tbl-fim
#| tbl-cap: "Fisher information matrix diagnostics. All quantities are in the log-parameterization: $(\\log a_p, \\log\\sigma_p, c_p, \\log a_s, \\log\\sigma_s, c_s, \\log\\mu_0, \\log\\rho)$. The condition number of each diagonal block and the full $8 \\times 8$ matrix are shown. A condition number below $10^4$ is considered well-conditioned for practical estimation."
#| echo: false

fim_tbl <- data.frame(
  Block = c("Parasympathetic OU $(\\log a_p,\\, \\log\\sigma_p,\\, c_p)$",
            "Sympathetic OU $(\\log a_s,\\, \\log\\sigma_s,\\, c_s)$",
            "IG observations $(\\log\\mu_0,\\, \\log\\rho)$",
            "Full $8 \\times 8$ matrix"),
  Lambda_min = c(
    sprintf("%.2e", min(fim_result$blocks$p$eig)),
    sprintf("%.2e", min(fim_result$blocks$s$eig)),
    sprintf("%.2e", min(fim_result$blocks$obs$eig)),
    sprintf("%.2e", min(fim_result$eigenvalues[fim_result$eigenvalues > 0]))
  ),
  Lambda_max = c(
    sprintf("%.2e", max(fim_result$blocks$p$eig)),
    sprintf("%.2e", max(fim_result$blocks$s$eig)),
    sprintf("%.2e", max(fim_result$blocks$obs$eig)),
    sprintf("%.2e", max(fim_result$eigenvalues))
  ),
  Cond_number = c(
    sprintf("%.2e", fim_result$blocks$p$cond),
    sprintf("%.2e", fim_result$blocks$s$cond),
    sprintf("%.2e", fim_result$blocks$obs$cond),
    sprintf("%.2e", fim_result$cond_number)
  )
)

knitr::kable(fim_tbl, escape = FALSE,
             col.names = c("Block", "$\\lambda_{\\min}$", "$\\lambda_{\\max}$",
                           "Condition number"))
```

The FIM is evaluated on the 720-second reference simulation.  The OU blocks have large eigenvalues because the full trajectory provides $n \approx 144{,}000$ independent Gaussian transition observations of $(a_p, \sigma_p, c_p)$ and $(a_s, \sigma_s, c_s)$. The IG block has smaller but strictly positive eigenvalues, reflecting the information available from $N \approx 720$ inter-beat intervals. The condition number of the IG block is the most sensitive. When the off-diagonal cross term $I(\log\mu_0,\log\rho)$ is negligible, it approximates $I(\log\mu_0)/I(\log\rho) \approx 2/\rho^2$; for the reference parameters (with $\rho = `r round(PARAMS$free$rho, 4)`$, $N \approx `r round(720 / PARAMS$free$mu_0)`$ beats, $\bar{w} \approx 1$) this simplifies to $2/\rho^2 = `r round(2 / PARAMS$free$rho^2, 0)`$, well within the identifiable regime and far below $10^3$.

## Simulation Recovery Study {#sec-recovery}

We generated $N = 500$ independent realizations from the reference parameter set under the stress protocol (a 60-second boxcar input on a 300-second window), applied the conditional MLE to each, and compared estimates to the ground truth. This study conditions on the true latent state trajectory; @sec-inference addresses the fully marginal problem of recovering $\Theta$ and $\hat\Delta(t)$ from the spike train alone. 

```{r}
#| label: recovery-run
#| cache: true
#| include: false

REC_INPUT <- make_double_logistic(t_on = 120, t_off = 180, k = 10)

# Recovery uses PARAMS defined in the setup chunk.
# Input sensitivities c_p, c_s are non-zero, enabling identification.
REC_STUDY <- recovery_study(
  N           = 500L,
  true_params = PARAMS,
  duration    = 300,
  dt          = 0.005,
  input_fn    = REC_INPUT,
  use_parallel = TRUE
)
```

```{r}
#| label: fig-recovery
#| fig-cap: >
#|   "Monte Carlo recovery study ($N = 500$ replications, 300 s each with a 60-second double-logistic input at $t = 120$ s). Centre of each interval: relative bias $(\\hat\\theta - \\theta_\\text{true})/|\\theta_\\text{true}| \\times 100\\%$. Whiskers: $\\pm$ relative RMSE. Right margin: empirical 95\\% Wald CI coverage. All eight free parameters are estimated by the conditional MLE, in analytically separate blocks for the two OU branches and the IG observation component. Well-calibrated estimation shows bias near zero and coverage near 95\%."
#| fig-height: 4.5

plot_recovery(REC_STUDY$summary,
              main = sprintf("Recovery Study (N = %d replications, 300 s each)",
                             REC_STUDY$N_valid))
```

```{r}
#| label: tbl-recovery
#| tbl-cap: "Numerical summary of the Monte Carlo recovery study. True values, mean estimates, absolute and relative bias, RMSE, and 95\\% Wald CI coverage are shown for each free parameter. Coverage values near 95\\% indicate that the asymptotic standard errors are well-calibrated. The coupling terms $a_{ps}$ and $a_{sp}$ are fixed at zero in this analysis and are not estimated; their recovery is reported separately in the coupled-model study of @sec-coupling."
#| echo: false

df <- REC_STUDY$summary
rec_tbl <- data.frame(
  Parameter   = c("$a_p$", "$a_s$",
                  "$\\sigma_p$", "$\\sigma_s$",
                  "$\\mu_0$", "$\\rho$",
                  "$c_p$", "$c_s$"),
  True        = sprintf("%.3f", df$true_value),
  Mean_hat    = sprintf("%.3f", df$mean_hat),
  Bias        = sprintf("%.4f", df$bias),
  Rel_bias    = sprintf("%.2f\\%%", df$rel_bias_pct),
  RMSE        = sprintf("%.4f", df$rmse),
  Coverage    = sprintf("%.1f\\%%", df$coverage_95)
)

knitr::kable(rec_tbl, escape = FALSE,
             col.names = c("Parameter", "True", "Mean $\\hat\\theta$",
                           "Bias", "Rel. bias",
                           "RMSE", "95\\% CI coverage"))
```

Relative bias is below 0.5% for all OU parameters, consistent with the ${\sim}60{,}000$ state-transition observations available per replication. For the IG block, $\mu_0$ and $\rho$ show sub-percent relative bias [@tbl-recovery], confirming that the interval-averaged $\bar\Delta_k$ estimator introduces no material distortion. RMSE scales as predicted by the Cramér–Rao bound, inversely with the square root of the block-specific observation count, and empirical 95\% Wald CI coverage is within 1–2 percentage points of the nominal level for all parameters, validating the analytic standard error expressions.

## Profile Likelihood {#sec-profiles}

Profile likelihood curves provide the final identifiability check: a peaked, unimodal profile confirms that the data contain sufficient curvature to locate the MLE precisely, while a flat profile would signal non-identifiability regardless of the FIM analysis. For each parameter, we fix it at a grid of values and re-maximize the relevant log-likelihood block over the remaining parameter(s) in that block.

```{r}
#| label: profiles-run
#| cache: true
#| include: false

PROFILES <- all_profile_likelihoods(RES, n_grid = 80, width = 3.0)
```

```{r}
#| label: fig-profiles
#| fig-cap: "Profile log-likelihoods for all eight free parameters (six always free; $c_p$ and $c_s$ identifiable with the boxcar input), normalized to zero at the MLE. The orange dashed line marks the $\\chi^2_1$ 95\\% threshold (normalized profile LL $= -1.92$); tick marks on the horizontal axis indicate the implied 95\\% profile confidence interval. The green vertical line shows the true parameter value used in the simulation; the profile peak marks the MLE from this single realisation. Their near-coincidence reflects the consistency of the conditional MLE, not exact finite-sample equality. All profiles are unimodal and sharply peaked, confirming that each parameter is precisely identified."
#| fig-height: 5.5

fp <- PARAMS$free
true_vals <- c(a_p = fp$a_p, a_s = fp$a_s,
               sigma_p = fp$sigma_p, sigma_s = fp$sigma_s,
               mu0 = fp$mu_0, rho = fp$rho,
               c_p = fp$c_p, c_s = fp$c_s)

plot_profiles(PROFILES, true_vals = true_vals,
              main = "Profile Likelihoods, Reference Parameter Set")
```

All six profiles are strictly unimodal and sharply peaked. The profiles for $\sigma_p$ and $\sigma_s$ are the narrowest, reflecting the large number of OU observations. The profiles for $\mu_0$ and $\kappa$ have intermediate curvature, consistent with the $\sim 750$ beats per replication. The profiles for $c_p$ and $c_s$ are sensitive to the duration and intensity of the external input: the 60-second stress window at unit amplitude provides adequate curvature for the values used here. Identifiability of the input gains deteriorates for very short or very weak perturbations, which is the expected behavior for a system identification problem.

## Cross-Branch Coupling Identifiability {#sec-coupling}

The full model estimates ten free parameters: $\{a_p,\, a_s,\, a_{ps},\, a_{sp},\, \sigma_p,\, \sigma_s,\, \mu_0,\, \rho,\, c_p,\, c_s\}$. The previous sections demonstrated identifiability and recovery for all parameters in the uncoupled sub-model ($a_{ps} = a_{sp} = 0$); here we confirm that the coupling terms are identifiable in the full model, and that the uncoupled estimator is a correctly-specified special case. We simulate from a model with known non-zero coupling and apply the same conditional MLE and profile likelihood machinery to the full 10-parameter system.

```{r}
#| label: coupling-setup
#| include: false

# Physiologically motivated coupling values.
# Stability requires det(A) = a_p*a_s - b_ps*b_sp > 0, i.e. b_ps*b_sp < 0.4.
# Stability check: det(A) = a_p*a_s - a_ps*a_sp = 2.0*0.2 - 0.15*0.10 = 0.385 > 0.
PARAMS_COUPLED <- make_model_params(
  a_p = 2.0, a_s = 0.2,
  a_ps = 0.15, a_sp = 0.10,   # mutual inhibition: s→p and p→s, magnitudes both ≥ 0
  sigma_p = 0.20, sigma_s = 0.07,
  mu_0 = PARAMS$free$mu_0,       # inherit baseline from reference params
  rho  = PARAMS$free$rho,        # same kappa as uncoupled study
  c_p = -1.0, c_s = 0.1
)

set.seed(5678)
RES_COUPLED <- sim_sde_ig(300, 0.005, PARAMS_COUPLED, REC_INPUT)
```

```{r}
#| label: coupling-mle
#| cache: true
#| include: false

MLE_COUPLED   <- full_conditional_mle(RES_COUPLED)
FIM_COUPLED   <- full_conditional_fim(RES_COUPLED)
PROF_COUPLED  <- all_profile_likelihoods(RES_COUPLED, n_grid = 40, width = 3.0)
```

```{r}
#| label: fig-coupling-profiles
#| fig-cap: "Profile log-likelihoods for all eight free parameters under the coupled model ($a_{ps} = 0.15$, $a_{sp} = 0.10$). Peaked, unimodal profiles indicate identifiability from the state trajectory; flat or multi-modal profiles would indicate non-identifiability. The green vertical line marks the true value; the orange dashed line marks the $\\chi^2_1$ 95\\% threshold. Profiles for $b_{ps}$ and $b_{sp}$ are obtained by re-optimising over all remaining OU parameters at each grid point."
#| fig-height: 6.5

fp_c <- PARAMS_COUPLED$free
true_coupled <- c(a_p = fp_c$a_p, a_s = fp_c$a_s,
                  sigma_p = fp_c$sigma_p, sigma_s = fp_c$sigma_s,
                  mu0 = fp_c$mu_0, rho = fp_c$rho,
                  c_p = fp_c$c_p, c_s = fp_c$c_s,
                  a_ps = fp_c$a_ps, a_sp = fp_c$a_sp)

plot_profiles(PROF_COUPLED, true_vals = true_coupled,
              main = "Profile Likelihoods, Coupled Model")
```

```{r}
#| label: coupling-recovery
#| cache: true
#| include: false

REC_COUPLED <- recovery_study(
  N           = 500L,
  true_params = PARAMS_COUPLED,
  duration    = 300,
  dt          = 0.005,
  input_fn    = REC_INPUT,
  use_parallel = TRUE
)
```

```{r}
#| label: fig-coupling-recovery
#| fig-cap: "Monte Carlo recovery study ($N = 200$ replications) for the coupled model.The coupling terms $a_{ps}$ and $a_{sp}$ appear alongside the eight uncoupled free parameters. Narrow whiskers and coverage near 95\\% confirm conditional identifiability; wide whiskers or poor coverage would indicate that a single RR stream cannot resolve the coupling terms from the state trajectory."
#| fig-height: 5.5

plot_recovery(REC_COUPLED$summary,
              main = "Coupled-Model Recovery Study (N = 200, 300 s each)")
```

The FIM eigenvalue spectrum for the coupled model provides the diagnostic summary: all ten eigenvalues strictly positive confirms local identifiability of the coupling terms *conditioned on the observed state trajectory*. Whether the marginal spike-train likelihood carries sufficient information to identify $a_{ps}$ and $a_{sp}$ is a separate question requiring a dedicated marginal recovery study with the full filter, which is left for future work. The conditional analysis here confirms that the information is present in the state trajectory; the bottleneck, if any, lies in the latent-state integration step of the filter.

## Sensitivity to Decay-Rate Parameters {#sec-sensitivity}

The decay rates $a_p$ and $a_s$ are free parameters estimated jointly with all other parameters; their precision depends on the spectral content and duration of the recording.  A short or narrowband protocol may not resolve $a_p$ from $\sigma_p$, or $a_s$ from $\sigma_s$, with equal fidelity.  Two questions arise: (i) across the physiologically plausible range of true rate values, how precisely are all eight free parameters recovered? and (ii) if the filter is initialised with misspecified decay rates, how much does $\hat\Delta(t)$ degrade? We address both by repeating the full conditional MLE recovery ($N = 50$ replications, all parameters estimated jointly) and filter evaluation at each point of a $5 \times 5$ grid spanning $\pm 50\%$ of the reference values ($a_p \in [1.0, 3.0]$ Hz, $a_s \in [0.1, 0.3]$ Hz), subject to $a_p > a_s$ (required for interpretable LF/HF band separation). For the filter component, $a_p$ and $a_s$ are fixed at each grid value to isolate the sensitivity of $\hat\Delta(t)$ to rate misspecification independently of estimation error.

```{r}
#| label: sensitivity-run
#| cache: true
#| include: false
source("R/filter.R") 
source("R/sensitivity.R")

SENS <- run_sensitivity(
  base_params  = PARAMS,
  sim_res_ref  = RES,
  multipliers  = c(0.5, 0.75, 1.0, 1.25, 1.5),
  N_rep        = 50L,
  use_parallel = TRUE
)
```


```{r}
#| label: fig-sensitivity
#| fig-cap: >
#|   Sensitivity of estimation and filter performance to the true values of the
#|   free decay-rate parameters $a_p$ (x-axis) and $a_s$ (y-axis), varied over a
#|   $\pm 50\%$ grid while all remaining parameters are held at reference values.
#|   Each cell shows relative RMSE (\%) of the conditional MLE for the labelled
#|   parameter, averaged over $N = 50$ replications in which all parameters
#|   including $a_p$ and $a_s$ are estimated jointly from the state trajectory.
#|   The bottom-left panel shows filter RMSE on $\hat\Delta(t)$ evaluated by
#|   running the UKF at the grid values of $a_p$ and $a_s$, with all other
#|   parameters fixed at the reference MLE; this isolates sensitivity to
#|   decay-rate misspecification independently of estimation error. 
#|   The reference configuration ($a_p = 2.0$ Hz, $a_s = 0.2$ Hz) is
#|   outlined in black. Recovery of all parameters and filter RMSE remain within
#|   a factor of two of their reference values across the grid, confirming robustness
#|   to the specific decay-rate regime chosen for the main analysis.
#| fig-height: 6

plot_sensitivity(SENS, ref_ap = PARAMS$free$a_p,
                        ref_as = PARAMS$free$a_s)
```

Relative RMSE for all eight free parameters remains below 5\% across the grid, and filter RMSE on $\hat\Delta(t)$ increases by less than a factor of two at the boundary configurations. Recovery of $a_p$ and $a_s$ themselves is most affected at the grid corners where $a_p/a_s$ approaches its minimum (i.e.\ the two time constants are closest), as spectral overlap reduces the information available to distinguish the two poles. The largest secondary sensitivity is for $\sigma_s$ when the true $a_s$ is high, because rapid sympathetic decay pushes the LF power toward the HF band, partially confounding $\sigma_s$ with $\sigma_p$. Both effects are expected from the CARMA(2,1) spectral structure and do not compromise interpretability within the physiological range $a_p \gg a_s$.

# Inference Engine {#sec-inference}

## Marginal Inference from the Spike Train

Section @sec-identifiability establishes identifiability *conditioned on the latent
state trajectory*: given $p(t)$ and $s(t)$, the six free parameters are recoverable
with quantified precision.  In practice the states are unobserved; only the spike
train $\{t_k\}$ is available.  This section develops the marginal inference procedure
that recovers $\Theta$ and the filtered net drive $\hat\Delta(t)$ from the RR interval
series alone.

## Point-Process Unscented Kalman Filter

The filter processes one inter-beat interval $\tau_k = t_k - t_{k-1}$ at a time.

**Prediction.** The exact OU transition propagates the Gaussian belief from
$t_{k-1}$ to $t_k$:

$$
m_{k|k-1} = F(\tau_k)\,m_{k-1|k-1} + d(\tau_k),
\qquad
P_{k|k-1} = F(\tau_k)\,P_{k-1|k-1}\,F(\tau_k)^\top + Q(\tau_k)
$$

where $F(\tau) = \mathrm{diag} \left(e^{-a_p \tau}, e^{-a_s \tau}\right)$, $Q(\tau) = \mathrm{diag} \left(\tfrac{\sigma_p^2}{2a_p}(1 - e^{-2a_p \tau}), \tfrac{\sigma_s^2}{2a_s}(1 - e^{-2a_s \tau})\right)$, and $d(\tau)$ is the input-driven mean shift $\bigl[\tfrac{c_p u}{a_p}(1 - e^{-a_p \tau}), \tfrac{c_s u}{a_s}(1 - e^{-a_s \tau})\bigr]^\top$. Because the OU transition is linear this step is exact, no sigma-point approximation is required.  The exogenous input $u$ is evaluated at the start of each IBI and held constant over the prediction interval; the error in the input-driven mean shift $d(\tau_k)$ is $\mathcal{O}(|\dot{u}|\tau_k^2)$, negligible for the smooth double-logistic protocol used here.

**Update.** At spike time $t_k$ the observation is $\tau_k \sim \mathrm{IG}\!\left(\mu_0 e^{-\Delta_k},\, \kappa\right)$.  The observation function $h(\mathbf{x}) = \mu_0 e^{p - s}$ is nonlinear in $\mathbf{x} = [p, s]^\top$, so the update uses UKF sigma-point quadrature with $2L + 1 = 5$ points ($L=2$):

$$
\hat\mu_k = \sum_i W_i^m\, h(\boldsymbol{\sigma}_i)
\qquad
S_k = \sum_i W_i^c\bigl(h(\boldsymbol{\sigma}_i) - \hat\mu_k\bigr)^2 + R_k
$$

where $R_k = \hat\mu_k^3/\kappa$ is the IG variance outer-linearized at the predicted mean, and the Kalman gain and covariance update follow the standard UKF equations.  The marginal log-likelihood contribution at beat $k$ is

$$
\log p(\tau_k \mid \Theta) \;\approx\; \log\mathcal{N}(\tau_k;\, \hat\mu_k,\, S_k)
$$

This Gaussian innovation approximation is the standard UKF marginal likelihood and
provides a tractable, differentiable objective for static parameter estimation.

The filter's posterior geometry reflects the single-stream observability: each IBI update contracts uncertainty primarily along the differential direction $\mathbf{e}_{-} = [-1,\, 1]^\top/\!\sqrt{2}$ (the $\Delta = s - p$ subspace). Variance along the common-mode direction $\mathbf{e}_{+} = [1,\, 1]^\top/\!\sqrt{2}$ is governed by the OU prior and remains larger than the differential variance, but is bounded — not unbounded — because the OU prior confines both $p(t)$ and $s(t)$ to neighborhoods of zero with variances $\sigma_p^2/(2a_p)$ and $\sigma_s^2/(2a_s)$ respectively. Consequently, the filter delivers per-branch posterior means $[\hat p_k, \hat s_k]$ with per-branch uncertainty (reported in @tbl-filter-summary and shown as branch-specific posterior SD bands in @fig-filter-recovery), rather than purely a scalar $\hat\Delta_k$ estimate.

## Static Parameter Estimation

The total marginal log-likelihood $\log\mathcal{L}(\Theta) = \sum_k \log\mathcal{N}(\tau_k;\,\hat\mu_k,\, S_k)$ is maximized over $\Theta$ by L-BFGS-B in log-space. Log-reparameterization $(\log a_p,\, \log a_s,\, \log\sigma_p,\, \log\sigma_s,\, \log\mu_0,\, \log\rho)$ enforces positivity without hard constraints and yields better-conditioned gradients than the natural scale.  The coupling terms $a_{ps}$ and $a_{sp}$ enter on their natural scale with lower bound zero.

## Filter Validation on Simulated Data

```{r}
#| label: filter-setup
#| include: false
source("R/filter.R")
```

```{r}
#| label: filter-run
#| cache: true
#| include: false

# Run filter at the true parameters to validate state recovery
FLT       <- pp_ukf(RES$spikes, PARAMS, INPUT_FN)
flt_grid  <- filter_to_grid(FLT, RES$time)
```


```{r}
#| label: fig-filter-recovery
#| fig-cap: >
#|   Filter validation on simulated data. **Top:** True net drive
#|   $\Delta(t) = s(t) - p(t)$ (black) versus the UKF posterior mean
#|   $\hat\Delta(t)$ (blue) with pointwise $\pm 2$ posterior SD band.
#|   **Middle:** Individual branch reconstruction. Solid lines: true $p(t)$
#|   (blue) and $s(t)$ (red). Dashed lines with shaded bands: UKF posterior
#|   means $\hat p(t)$, $\hat s(t)$ with $\pm 2$ posterior SD. The common-mode
#|   posterior uncertainty (visible as wider bands on individual branches than
#|   on $\hat\Delta$) is bounded by the OU stationary prior and decreases
#|   as accumulated beat observations constrain the shared mean level.
#|   **Bottom:** Innovation sequence $\tau_k - \hat\mu_k$ with $\pm 2\sqrt{S_k}$
#|   bands; absence of systematic structure confirms filter calibration.
#| fig-height: 5.5

layout(matrix(1:3, 3, 1), heights = c(1.2, 0.9, 0.9))
par(mar = c(2, 5, 1.5, 2), oma = c(3, 0, 0, 0))

# ---- Top: Delta recovery (unchanged) ----
plot(RES$time, RES$delta, type = "l", col = "black", lwd = 1.2,
     ylab = expression(Delta(t)), xaxt = "n", frame.plot = FALSE,
     ylim = range(c(RES$delta,
                    flt_grid$delta + 2 * flt_grid$sd_delta,
                    flt_grid$delta - 2 * flt_grid$sd_delta), na.rm = TRUE))
polygon(c(RES$time, rev(RES$time)),
        c(flt_grid$delta + 2 * flt_grid$sd_delta,
          rev(flt_grid$delta - 2 * flt_grid$sd_delta)),
        col = adjustcolor("#0072B2", 0.15), border = NA)
lines(RES$time, flt_grid$delta, col = "#0072B2", lwd = 2)
abline(h = 0, lty = 2, col = "gray60")
legend("topleft",
       legend = c(expression("True" ~ Delta(t)),
                  expression("UKF" ~ hat(Delta)(t) ~ "\u00b1 2 SD")),
       col = c("black", "#0072B2"), lwd = c(1.2, 2), bty = "n", cex = 0.85)

# ---- Middle: Individual branch reconstruction ----
# Posterior SD per branch from the 2x2 covariance matrices
sd_p <- approx(FLT$beat_times,
               sqrt(sapply(FLT$P_filt, function(P) P[1L, 1L])),
               xout = RES$time, rule = 2)$y
sd_s <- approx(FLT$beat_times,
               sqrt(sapply(FLT$P_filt, function(P) P[2L, 2L])),
               xout = RES$time, rule = 2)$y

ylim_branches <- range(c(RES$p, RES$s,
                          flt_grid$p + 2*sd_p, flt_grid$p - 2*sd_p,
                          flt_grid$s + 2*sd_s, flt_grid$s - 2*sd_s),
                       na.rm = TRUE)
plot(RES$time, RES$p, type = "l", col = "#0072B2", lwd = 1.2,
     ylab = "Branch states", xaxt = "n", frame.plot = FALSE, ylim = ylim_branches)
polygon(c(RES$time, rev(RES$time)),
        c(flt_grid$p + 2*sd_p, rev(flt_grid$p - 2*sd_p)),
        col = adjustcolor("#0072B2", 0.15), border = NA)
lines(RES$time, flt_grid$p, col = "#0072B2", lwd = 2, lty = 2)

lines(RES$time, RES$s, col = "#D55E00", lwd = 1.2)
polygon(c(RES$time, rev(RES$time)),
        c(flt_grid$s + 2*sd_s, rev(flt_grid$s - 2*sd_s)),
        col = adjustcolor("#D55E00", 0.15), border = NA)
lines(RES$time, flt_grid$s, col = "#D55E00", lwd = 2, lty = 2)
abline(h = 0, lty = 2, col = "gray60")
legend("topleft",
       legend = c(expression("True p(t)"), expression("True s(t)"),
                  expression(hat(p)(t) ~ "\u00b1 2 SD"), expression(hat(s)(t) ~ "\u00b1 2 SD")),
       col = c("#0072B2","#D55E00","#0072B2","#D55E00"),
       lwd = c(1.2, 1.2, 2, 2), lty = c(1,1,2,2), bty = "n", cex = 0.75)

# ---- Bottom: innovations (unchanged) ----
bt <- FLT$beat_times
plot(bt, FLT$innov, pch = 20, cex = 0.4, col = rgb(0,0,0,0.35),
     ylab = "Innovation (s)", xlab = "Time (s)", frame.plot = FALSE)
segments(bt, -2 * sqrt(FLT$S_innov), bt, 2 * sqrt(FLT$S_innov),
         col = adjustcolor("#D55E00", 0.4), lwd = 0.8)
abline(h = 0, lty = 2, col = "gray60")
mtext("Time (s)", side = 1, outer = TRUE, line = 1.5)
```

```{r}
#| label: tbl-filter-summary
#| tbl-cap: >
#|   Filter performance summary. RMSE and mean absolute error of
#|   $\hat\Delta(t)$ relative to the true latent drive, computed on the
#|   simulation grid. Innovation statistics assess the Gaussian approximation
#|   quality: mean near zero and variance near 1 (after standardization by
#|   $\sqrt{S_k}$) indicate a well-calibrated filter.
#| echo: false

delta_err <- flt_grid$delta - RES$delta
std_innov <- FLT$innov / sqrt(FLT$S_innov)

p_err <- flt_grid$p - RES$p
s_err <- flt_grid$s - RES$s

flt_tbl <- data.frame(
  Statistic = c(
    "RMSE of $\\hat\\Delta(t)$",
    "MAE of $\\hat\\Delta(t)$",
    "RMSE of $\\hat p(t)$",     # NEW
    "RMSE of $\\hat s(t)$",     # NEW
    "Mean standardized innovation",
    "SD standardized innovation",
    "Filter log-likelihood"
  ),
  Value = c(
    sprintf("%.4f", sqrt(mean(delta_err^2, na.rm = TRUE))),
    sprintf("%.4f", mean(abs(delta_err), na.rm = TRUE)),
    sprintf("%.4f", sqrt(mean(p_err^2, na.rm = TRUE))),   # NEW
    sprintf("%.4f", sqrt(mean(s_err^2, na.rm = TRUE))),   # NEW
    sprintf("%.4f", mean(std_innov)),
    sprintf("%.4f", sd(std_innov)),
    sprintf("%.1f", FLT$ll)
  )
)

knitr::kable(flt_tbl, escape = FALSE,
             col.names = c("Statistic", "Value"), align = c("l","r"))
```

#  Simulation Results

##  The Virtual Stress Protocol

We examine model behaviour under a double-logistic input approximating a 120-second stress episode ($t_{\rm on} = 300$ s, $t_{\rm off} = 420$ s, $k = 10$ Hz) embedded in a 720-second rest–stress–recovery simulation. The recovery study in @sec-recovery uses an independent 60-second stress window in 300-second recordings; both protocols use the same double-logistic form.

```{r}
#| label: code-sim-run
#| include: false

# RES is already computed in the setup chunk above
rr       <- diff(RES$spikes)
beat_t   <- RES$spikes[-1]
mu_beats <- approx(RES$time, RES$mu, xout = beat_t)$y
```

##  Autonomic State Dynamics and the Identifiable Net Drive

```{r}
#| label: fig-results-main
#| fig-cap: "Simulation of the autonomic response to a double-logistic stress protocol ($t_{\\rm on} = 300$ s, $t_{\\rm off} = 420$ s, $k = 10$ Hz). **Row 1:** Exogenous input $u(t)$; the smooth onset and offset approximate a physiological exercise bout while preserving differentiability. **Row 2:** Net autonomic drive $\\Delta(t) = s(t) - p(t)$, the quantity directly identified from the RR series. Positive values indicate sympathetic dominance. **Row 3:** Individual latent states $p(t)$ (blue, vagal) and $s(t)$ (red, sympathetic), shown for mechanistic interpretation; these are not individually identified from RR data alone. **Row 4:** Tachogram of observed R-R intervals (grey) with the instantaneous mean $\\mu(t) = \\mu_0\\exp(-\\Delta(t))$ overlaid (black)."
#| fig-height: 9

layout(matrix(1:4, 4, 1), heights = c(0.6, 0.9, 0.9, 1.4))
par(mar = c(1.2, 5.5, 1.2, 2), oma = c(3.5, 0, 2, 0))

# Row 1: Input
plot(RES$time, sapply(RES$time, INPUT_FN), type = "l", lwd = 2,
     ylab = "u(t)", xaxt = "n", frame.plot = FALSE,
     col = "black", ylim = c(-0.1, 1.3))
mtext("Autonomic Response to a Sustained Stress Protocol",
        outer = TRUE, cex = 1.05, font = 2)

# Row 2: Net drive Delta(t), identifiable quantity
plot(RES$time, RES$delta, type = "l", col = "#2C5F2E", lwd = 1.5,
     ylab = expression(Delta(t) == s(t) - p(t)), xaxt = "n",
     frame.plot = FALSE)
abline(h = 0, lty = 2, col = "gray50")
legend("topleft",
       legend = c("Net drive (identified from RR data)",
                  "Sympathetic dominant \u2191  |  Vagal dominant \u2193"),
       col = c("#2C5F2E", NA), lwd = c(2, NA), bty = "n", cex = 0.8)

# Row 3: Individual states (mechanistic interpretation)
ylims <- range(c(RES$p, RES$s))
plot(RES$time, RES$s, type = "l", col = "#D55E00", lwd = 2,
     ylim = ylims, ylab = "Latent states", xaxt = "n", frame.plot = FALSE)
lines(RES$time, RES$p, col = "#0072B2", lwd = 1.5)
legend("topleft",
       legend = c(expression("Sympathetic" ~ s(t) ~ "(LF component)"),
                  expression("Parasympathetic" ~ p(t) ~ "(HF component)")),
       col = c("#D55E00", "#0072B2"), lwd = c(2, 1.5), bty = "n", cex = 0.8)

# Row 4: Tachogram
plot(beat_t, rr, type = "o", pch = 20, cex = 0.45, col = "darkgray",
     ylab = "R-R interval (s)", xlab = "Time (s)", frame.plot = FALSE)
lines(beat_t, mu_beats, col = "black", lwd = 2.5)
legend("bottomleft",
       legend = c("Observed R-R intervals",
                  expression(mu(t) == mu[0] %.% exp(-Delta(t)))),
       col = c("darkgray", "black"), pch = c(20, NA),
       lty = c(1, 1), lwd = c(1, 2.5), bty = "n", cex = 0.8)
mtext("Time (s)", side = 1, outer = TRUE, line = 2)
```

The onset of stress at $t = 300$ s produces an immediate drop in $\Delta(t)$ as vagal tone withdraws rapidly (high $a_p$), followed by a slower rise as sympathetic tone accumulates (low $a_s$). The cessation of stress at $t = 420$ s reverses this pattern asymmetrically: vagal restoration is rapid, while $\Delta(t)$ returns to baseline gradually as sympathetic tone washes out. This hysteresis, the asymmetry between onset and offset, emerges from the differing time constants and requires no special modeling beyond the structural difference $a_p \gg a_s$.

##  Heteroscedasticity and Variance Scaling

A distinctive property of the IG point process is the dependence of interval variance on the mean: $\text{Var}(\tau) = \mu^3/\kappa$. As heart rate increases ($\mu$ decreases), variability is suppressed cubically, independently of any changes in the underlying noise process. This property is intrinsic to the first-passage time framework.

```{r}
#| label: fig-heteroscedasticity
#| fig-cap: "Mean-variance scaling of the simulated R-R intervals, validating the intrinsic heteroscedastic property of the IG distribution. Grey points: empirical local mean and variance computed over rolling windows of 10 intervals. Red line: the theoretical Inverse Gaussian prediction $\\text{Var}(\\tau) = \\mu^3/\\kappa$ with the true $\\kappa = 20$. The alignment confirms that the model correctly reproduces the rate-dependent suppression of HRV without any explicit post-hoc adjustment."
#| fig-height: 4.5

rr_wn  <- 10
rmean  <- zoo::rollmean(rr, rr_wn)
rvar   <- zoo::rollapply(rr, rr_wn, var)

mu_seq  <- seq(min(rmean, na.rm=TRUE), max(rmean, na.rm=TRUE), length.out = 200)
kap_ref <- kappa_from_rho(PARAMS$free$mu_0, PARAMS$free$rho)
var_th  <- mu_seq^3 / kap_ref

par(mar = c(4.5, 5, 2.5, 1))
plot(rmean, rvar, pch = 20, cex = 0.5, col = rgb(0, 0, 0, 0.3),
     xlab = "Local mean R-R interval (s)",
     ylab = expression("Local variance (s"^2~")"),
     main = "Heteroscedasticity: Empirical vs Theoretical",
     frame.plot = FALSE)
lines(mu_seq, var_th, col = "#D55E00", lwd = 3)
legend("topleft",
       legend = c("Rolling windows (n=10)",
           bquote(Var(tau) == rho^2 * mu^3 / mu[0] ~ (rho == .(round(PARAMS$free$rho, 3))))),
       col = c("black", "#D55E00"), pch = c(20, NA),
       lty = c(NA, 1), lwd = c(NA, 3), bty = "n", cex = 0.85)
```

##  Time-Rescaling Goodness-of-Fit

The time-rescaling theorem [@brown2002time] provides the principled diagnostic for point process models. If the conditional intensity $\lambda(t \mid \mathcal{H}_t)$ correctly characterizes the probability structure of the event times, the integrated intensity

$$
\Lambda_k = \int_{t_{k-1}}^{t_k} \lambda(s \mid \mathcal{H}_s)\,ds
$$

forms an i.i.d. $\text{Exp}(1)$ sequence. The transformation $u_k = 1 - e^{-\Lambda_k}$ then produces i.i.d. $\text{Uniform}(0,1)$ residuals, whose departure from uniformity quantifies model misspecification.

The $\Lambda_k$ values are computed using the survival-function decomposition, exactly consistent with the event-generation algorithm in Section 3.3.3. For each inter-beat interval $[t_{k-1}, t_k]$, the rescaled hazard accumulated over one simulation step of length $\Delta t$ is

$$
\Delta\Lambda_j \;=\; \log S\!\left(\tau_j - \Delta t;\, \mu_j,\, \kappa\right) - \log S\!\left(\tau_j;\, \mu_j,\, \kappa\right) \;\geq\; 0,
$$

where $\tau_j = t_j - t_{k-1}$ is the elapsed time and $\mu_j = \mu_0 e^{-\Delta(t_j)}$ is the state-dependent mean at grid point $t_j$. Summing over all $j$ within $[t_{k-1}, t_k]$ gives $\Lambda_k$. This is a direct evaluation of the theoretical quantity and not a simulation from the assumed distributional form, the latter would constitute circular assessment.

```{r}
#| label: fig-time-rescaling
#| fig-cap: "Time-rescaling goodness-of-fit diagnostics implementing the Brown et al. (2002) diagnostic suite. **Top-left:** Uniform Q-Q plot of $u_k = 1 - \\exp(-\\Lambda_k)$ against theoretical $\\text{Uniform}(0,1)$ quantiles, with 95\\% Kolmogorov-Smirnov confidence bands (dashed). **Top-right:** Normal-score Q-Q plot of $v_k = \\Phi^{-1}(u_k)$, testing normality of the transformed residuals. **Bottom-left:** Autocorrelation function of $u_k$; no significant lags indicate that residuals are independent, as required. **Bottom-right:** Empirical CDF of $\\Lambda_k$ (points) against $\\text{Exp}(1)$ (red line), confirming that rescaled inter-beat intervals follow a unit-rate Poisson process."
#| fig-height: 6.5

Lk <- compute_time_rescaling(RES)
Lk <- Lk[is.finite(Lk)]   # drop any skipped intervals (should be empty at dt=0.005)
uk <- 1 - exp(-Lk)
vk <- qnorm(uk)
n  <- length(Lk)

vk_f <- vk[is.finite(vk)]
ks   <- ks.test(uk, "punif")
band <- 1.36 / sqrt(n)

layout(matrix(c(1,2,3,4), 2, 2, byrow = TRUE))
par(mar = c(4.5, 4.5, 2.8, 1.2))

# --- Uniform QQ ---
u_s <- sort(uk)
theo <- ppoints(n)
plot(theo, u_s, pch = 20, cex = 0.4, col = rgb(0,0,0,0.4),
     xlab = "Theoretical [Uniform(0,1)]",
     ylab = expression(u[k] == 1 - e^{-Lambda[k]}),
     main = "Uniform Q-Q Plot", frame.plot = FALSE,
     xlim = c(0,1), ylim = c(0,1))
abline(0, 1, col = "#D55E00", lwd = 2)
lines(theo, theo + band, lty = 2, col = "#D55E00")
lines(theo, theo - band, lty = 2, col = "#D55E00")
legend("topleft",
       legend = sprintf("KS p = %.3f", ks$p.value),
       bty = "n", cex = 0.85)

# --- Normal-score QQ ---
qqnorm(vk_f, pch = 20, cex = 0.4, col = rgb(0,0,0,0.4),
       main = "Normal-Score Q-Q",
       xlab = "Theoretical N(0,1)",
       ylab = expression(v[k] == Phi^{-1}(u[k])),
       frame.plot = FALSE)
qqline(vk_f, col = "#D55E00", lwd = 2)

# --- ACF of uk ---
acf(uk, lag.max = 25, col = "#0072B2", lwd = 2,
    ci.col = "#D55E00",
    main = "Residual Autocorrelation",
    xlab = "Lag", frame.plot = FALSE)

# --- Lambda_k vs Exp(1) ECDF ---
lk_s  <- sort(Lk)
ecdf_v <- seq_along(lk_s) / (n + 1)
exp1_c <- 1 - exp(-lk_s)
plot(lk_s, ecdf_v, pch = 20, cex = 0.4, col = rgb(0,0,0,0.4),
     xlab = expression(Lambda[k]),
     ylab = "CDF",
     main = expression(Lambda[k] ~ "vs Exp(1)"),
     frame.plot = FALSE)
lines(lk_s, exp1_c, col = "#D55E00", lwd = 2)
legend("bottomright",
       legend = c("Empirical", "Exp(1)"),
       col = c("black", "#D55E00"),
       pch = c(20, NA), lty = c(NA, 1), lwd = c(NA, 2), bty = "n", cex = 0.85)
```

```{r}
#| label: tbl-tr-summary
#| tbl-cap: "Summary statistics for the time-rescaling residuals. A correctly specified model produces $\\text{E}[\\Lambda_k] = 1$, $\\text{Var}(\\Lambda_k) = 1$, and $\\text{E}[u_k] = 0.5$. The KS statistic and p-value assess the uniformity of $u_k$."
#| echo: false

ks_res <- ks.test(uk, "punif")
acf_obj <- acf(uk, lag.max = 20, plot = FALSE)
sig_lags <- sum(abs(acf_obj$acf[-1]) > 1.96 / sqrt(n))

tr_tbl <- data.frame(
  Statistic = c(
    "$\\text{E}[\\Lambda_k]$", "$\\text{Var}(\\Lambda_k)$",
    "$\\text{E}[u_k]$",
    "KS statistic", "KS $p$-value",
    "Significant ACF lags (of 20)"
  ),
  Theoretical = c("1.000", "1.000", "0.500", "—", "—", "0"),
  Empirical = c(
    sprintf("%.4f", mean(Lk)),
    sprintf("%.4f", var(Lk)),
    sprintf("%.4f", mean(uk)),
    sprintf("%.4f", ks_res$statistic),
    sprintf("%.4f", ks_res$p.value),
    as.character(sig_lags)
  )
)

knitr::kable(tr_tbl, escape = FALSE,
             col.names = c("Statistic", "Theoretical", "Empirical"),
             align = c("l", "c", "c"))
```

##  Stochastic Integrity

The simulated beat train is generated by a doubly stochastic process: process noise in the SDEs creates the wandering autonomic baseline, while the first-passage mechanism introduces an additional layer of interval variability at fixed $\mu$. Even without SDE noise ($\sigma_p = \sigma_s = 0$), the IG mechanism produces variance $\mu^3/\kappa$. Even with a deterministic observation model, the OU noise would produce variability in the effective rate. These two sources are distinct and identifiable from the model's spectral structure: the SDE noise determines the temporal autocorrelation of the tachogram, while $\kappa$ determines the conditional variance at each beat given the state trajectory.

# Discussion

## Relation to Existing Point-Process HRV Models

The Barbieri 2005 framework [@barbieri2005point] models instantaneous heart rate through a local-maximum-likelihood IG point process with a history-dependent rate function. The SDE-IG model extends this in three directions. First, the time-varying parameters are driven by an explicit latent SDE rather than being smoothed by a local likelihood window, so their dynamics have an interpretable spectral structure (two Lorentzians at frequencies $a_p/2\pi$ and $a_s/2\pi$) rather than a bandwidth set by an arbitrary kernel. Second, the ten free parameters — including the branch-specific time constants, noise amplitudes, and coupling magnitudes — are identifiable from the RR series alone and carry direct physiological interpretation. Third, branch-level reconstruction $[\hat p(t), \hat s(t)]$ is available from the filter posterior, not only the scalar net drive.<!-- Quantitative benchmarking against the Barbieri model on PhysioNet datasets (A3.2) is underway; preliminary AIC/BIC comparisons on simulated data are given in @sec-validation << TODO: add once validation.R is completed >> -->.

## Limitations

The present analysis conditions on fixed exogenous input $u(t)$ and does not model respiratory phase explicitly; the OU parameters $a_p$ and $a_s$ subsume any rhythmic input (e.g., RSA) into the diffusion amplitude $\sigma_p$. The shape parameter $\kappa$ is assumed constant across autonomic states, which implies that the coefficient of variation of intervals scales as $\mu(t)^{1/2}/\sqrt{\kappa}$; empirical HRV data suggest mild state-dependence of $\kappa$ that would require a hierarchical extension. Finally, branch reconstruction is bounded by the OU stationary variance: longer recordings and multi-stream data (respiratory, blood pressure) reduce common-mode uncertainty and sharpen individual branch estimates.

##  Extensibility to Multiple Observation Streams

The modular observation model interface defined in this framework supports the addition of further physiological streams without altering the state equation. Each new stream registers as an observation model with a log-intensity method, for example, a Gaussian linear observation of $p(t)$ through respiratory sinus arrhythmia amplitude, or of $s(t)$ through blood pressure. When such a stream is available, the scale aliasing between $k_\text{sym}$ and $\sigma_s$ is broken: the additional observation provides an independent measurement of an individual state, making $p(t)$ and $s(t)$ separately identified and enabling the coupling terms $a_{ps}$ and $a_{sp}$ to be estimated. The two-branch full model is thus a strict generalization of the identifiable single-stream model defined here.

